---
title: "DIN vs TN Scatterplot"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
source("C:/Users/andre/my_functions.R")

```

```{r, message = FALSE}
# Load the data
data <- read_csv("Data/processed_dataset.csv")

```

### Derive May-Oct averages

Yearly means for DIN and TN.

```{r, message=FALSE}
# Workflow:
# loop through each station and calculate a May-Oct average provided that there are at least 3 observations in a year.

vars <- c( "NITROGEN_TOTAL", "DIN")
stations <- unique(data$STATION_ID)

stationAvgs <- list()

for (station in stations) {
  
  station_avgs <- list()
  
  for (variable in vars) {
    # Filter for current station and include only non-NA values of current var
    data.sbst_station_variable <- data %>%
      filter(STATION_ID == station) %>%
      filter(!is.na(!!sym(variable))) # dynamically reference the variable

      # Determine which years have at least 3 obs/yr
      years_with_3_obs <- data.sbst_station_variable %>%
        group_by(Year) %>%
        summarise(n = sum(!is.na(!!sym(variable)))) %>%
        filter(n >= 3)

        data.sbst_station_variable <- data.sbst_station_variable %>%
          filter(Year %in% years_with_3_obs$Year) %>% # Only include years with >= 3 obs
          group_by(Year, STATION_ID) %>%
          summarise(MAY_OCT_AVG = mean(!!sym(variable), na.rm = TRUE))

        # Add the May-Oct average for variable to the station_avgs list
        station_avgs[[variable]] <- data.sbst_station_variable
      }
  # Only add station to the final list if it has valid averages for at least one variable
  if (length(station_avgs) > 0) {
    stationAvgs[[station]] <- station_avgs
  }
}

## List structure:
## list[[station]]$variable  variable = df with Year, STATION_ID, MAY_OCT_AVG


# Simplify the list
bound_averages <- list()
for (station in names(stationAvgs) ) {
  bound <- bind_rows(stationAvgs[[station]], .id = "variable")
  bound_averages[[station]] <- bound
}

# Combine all station averages into one dataframe
mayOctAvgs_df <- bind_rows(bound_averages)


nitro_data <- mayOctAvgs_df %>%
  pivot_wider(names_from = variable, values_from = MAY_OCT_AVG) %>%
  filter(!is.na(DIN) & !is.na(NITROGEN_TOTAL)) %>%
  arrange(STATION_ID, Year)

table(nitro_data$Year)
write_csv(nitro_data, "Data/DINvsTN.csv")

```

### Scatterplot

"Could you make a scatterplot of May-Oct mean DIN (x) vs May-Oct mean TN (y)?  Each data point would represent one station-year.  For most of our stations, the DIN data are pre-2000 and the TN data are post-2000 so I am not sure how many data points we will get, but I would like to show that most of the TN is in the form of DIN (i.e., the two should be correlated)."

```{r}
# Create the scatterplot
r_val <- cor(nitro_data$DIN, nitro_data$NITROGEN_TOTAL)

ggplot(nitro_data, aes(x = DIN, y = NITROGEN_TOTAL)) +
  geom_point(alpha = .65) +
  labs(title = "May-Oct Mean DIN vs TN",
       x = "DIN (mg/L)",
       y = "TN (mg/L)") +
  annotate("text", x = .22, y = 1.25, label = paste("r =", round(r_val, 3)),
           hjust = 1.1, vjust = 1.1, size = 4.5)
  theme_minimal()

  ggsave("figures/DINvsTN.jpg", width = 6, height = 6, dpi = 600)


```