---
title: "Reservoir Water Clarity: Long-Term Intrannual Trends"
output: html_document
toc: true  
toc_float: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
source("C:/Users/andre/my_functions.R")

```

```{r, message = FALSE}
# Load the data
data <- read_csv("Data/processed_dataset.csv")

```

# Visualizing Time series with Model Fits

From Paul: "I have come up with a list of 33 stations (below) for which I would like you to generate time series plots for variables that have at least 10 years of data inclusive of CHLa, Secchi, TN, DIN, TP and TSS.  In the output file, the plots should be organized by station (e.g., CHLa, TN and Secchi for station 1, then station 2, etc.), so I can compare trends.  Note that these plots should be based on the May-Oct average values and show the model line with slope and p statistics."


### Vectorize the station IDs

```{r}
# These stations have the longest record of Secchi data (since 1980), but were not included among trend sites. 
# ONLY Secchi trends are of interest at these 4 sites.
secchi_stations1 <- unlist(strsplit("2-APP020.23 2-LCN000.20 2-PKC001.84 2-SPE000.17 4ACRR008.32", " "))

# For all other sites, we are interested in trends for multiple variables.
# These stations have the best long-term records for Secchi with >25 years of data since ~1995.  These sites also have good records for CHLa, TN, TP, and in some cases, TSS.
secchi_stations2 <- unlist(strsplit("4ABSA000.62 4ABWR002.50 4ABWR010.55 4ABWR017.42 4ACCK001.80 4ACOA000.60 4AGIL002.39 4AROA158.22 4AROA163.76 4AROA167.34 4AROA175.63 4AROA180.21 4AROA183.64 4AROA192.94 4AROA196.05 4AWTH000.40 9-NEW087.14 9-NEW089.34 9-NEW092.66 9-NEW098.32 9-PKC000.00 9-PKC004.16", " "))

# These stations show the largest increasing or decreasing trends in CHLa (some not included as already in above vectors).
chlaTrends_stations <- c("4AROA038.49", "8-NAR047.69", "8-NAR054.17", "9-PKC004.16")

# These stations show the largest increasing or decreasing trends in Secchi (some not included as already in above vectors).
secchiTrends_stations <- c("1BCRO009.19", "4ASRE046.90")


# Group sites into vectors for use in loops throughout script
stations_of_interest_10yrs <- c(secchi_stations2, chlaTrends_stations, secchiTrends_stations)
all33_stations <- c(secchi_stations1, secchi_stations2, chlaTrends_stations, secchiTrends_stations)

# --------------------------

# Lake Anna stations
## These stations are not included in the 33 stations of interest. They are from the subsequent LACA data set that was obtained after originally identifying SoI and generating the time series plots.
laca.data <- openxlsx::read.xlsx("Data/Lake Anna Suppl.xlsx", sheet = 2)
laca.stations <- unique(laca.data$`DEQ.Site.#`)
laca.stations <- laca.stations[laca.stations %in% unique(data$STATION_ID)]  # Some LACA stations did not make into processed data (no station-year with 3 Nobs)

updated.SoI <- unique(c(stations_of_interest_10yrs, laca.stations)) # all stations of interest MINUS 5 secchi only stations
updated.SOI_COMPLETE <- unique(c(all33_stations, laca.stations)) # all stations of interest

# -----------------------------
# Subset data to only include stations where multiple variables are of interest. 
# Secchi-only station data can be drawn from the original `data` variable
data.sbst <- data %>% 
  filter(STATION_ID %in% updated.SoI) %>%
  select(-PHOSPHORUS_ORTHO, -TURBIDITY_HACH, TURBIDITY_NTU, AMMONIA_TOTAL, NITRATE_TOTAL)

```

### Derive May-Oct averages

Yearly means for each variable at each station iff that variable has >= 10 years of data.

```{r, message=FALSE}
# Workflow:
# loop through each station
# loop through each variable and determine if there are at least 10 unique years of data, and if so, determine for how many of those years there are at least 3 observations per year

# This vector and its specific order is important throughout this file; it is used to loop through the variables in the correct order
vars <- c("CHLa", "PHOSPHORUS_TOTAL", "NITROGEN_TOTAL", "DIN", "TSS", "SECCHI_DEPTH")

stationAvgs_10yrs <- list()

for (station in updated.SoI) {
  
  station_avgs <- list()
  
  for (variable in vars) {
    # Filter for current station and include only non-NA values of current var
    data.sbst_station_variable <- data.sbst %>%
      filter(STATION_ID == station) %>%
      filter(!is.na(!!sym(variable))) # dynamically reference the variable

    # Confirm at least 10 yrs of data
    if (length(unique(data.sbst_station_variable$Year)) >= 10) {
      # Determine which years have at least 3 obs/yr
      years_with_3_obs <- data.sbst_station_variable %>%
        group_by(Year) %>%
        summarise(n = sum(!is.na(!!sym(variable)))) %>%
        filter(n >= 3)

      if (nrow(years_with_3_obs) >= 10) {
        # Calculate May-Oct average for the current station and variable
        data.sbst_station_variable <- data.sbst_station_variable %>%
          filter(Year %in% years_with_3_obs$Year) %>% # Only include years with >= 3 obs
          group_by(Year, STATION_ID) %>%
          summarise(MAY_OCT_AVG = mean(!!sym(variable), na.rm = TRUE))

        # Add the May-Oct average for variable to the station_avgs list
        station_avgs[[variable]] <- data.sbst_station_variable
      }
    }
  }
  # Only add station to the final list if it has valid averages for at least one variable
  if (length(station_avgs) > 0) {
    stationAvgs_10yrs[[station]] <- station_avgs
  }
}


# ---------- Secchi Stations with < 10 yrs ------------------
# There are 5 stations for which we are allowing inclusion in trend estimates despite having < 10 years of data.
secchi_only <- data %>%
  filter(STATION_ID %in% secchi_stations1)

secchiOnly_avgs <- list()

# Loop through each station and identify which years have Nobs >= 3
for (station in secchi_stations1) {
  station_data <- secchi_only %>%
    filter(STATION_ID == station)
  
    # Determine which years have at least 3 obs/yr
      secchiYrs_3_obs <- station_data %>%
        group_by(Year) %>%
        summarise(n = sum(!is.na(SECCHI_DEPTH))) %>%
        filter(n >= 3)

    # Calculate May-Oct average for the current station and variable
        mayOct_avgs <- station_data %>%
          filter(Year %in% secchiYrs_3_obs$Year) %>% # Only include years with >= 3 obs
          group_by(Year, STATION_ID) %>%
          summarise(MAY_OCT_AVG = mean(SECCHI_DEPTH, na.rm = TRUE))
    
    secchiOnly_avgs[[station]] <- list(SECCHI_DEPTH = mayOct_avgs)
      
}

# Combine to make single list of all monthly averages
yearlyAvgs_complete <- c(stationAvgs_10yrs, secchiOnly_avgs)
## List structure:
## list[[station]]$variable  variable = df with Year, STATION_ID, MAY_OCT_AVG


```

### Run MBLM for each station and variable

Each station is stored as a list in `all_stations_avgs`. Each station list contains a list of variables with May-Oct averages for each year containing at least 3 observations of that variable.

#### Custom function to derive trends

```{r, warning = FALSE}
trends_func <- function(station, variable) {
  
  modelSummary <- list()  
  
  df <- yearlyAvgs_complete[[station]][[variable]]
  
  # Remove NA values -- there should not be any, but just in case.
  df <- df %>%
    filter(!is.na(MAY_OCT_AVG)) %>%
    ungroup()
  
  # Run mblm model
  model <- mblm::mblm(MAY_OCT_AVG ~ Year, data = df)
  mod.sum <- mblm::summary.mblm(model)
          
  # Store results (directly, without creating the combined station-variable name)
  modelSummary <- list(
    slope = mod.sum$coefficients[2, 1],
    MAD = mod.sum$coefficients["Year", "MAD"],
    pvalue = mod.sum$coefficients["Year", 4],
    intercept = mod.sum$coefficients[1, 1]
  )
  
  return(modelSummary)  # Return the results
}

# ------- Run custom function in for loop -------
# trends_results (outer list); contains -> list for each variable; contains -> lists for each station; which each contain -> 1 list with 4 elements (the model results)
## list[[station]]$variable  variable = list with slope, MAD, pval, intercept

# Trends for stations of interest/variables with 10+ yr and 3+ obs/yr
trends_results <- list()

for (station in updated.SOI_COMPLETE) {
  for (variable in vars) {
    if (variable %in% names(yearlyAvgs_complete[[station]])) {
  trends_results[[station]][[variable]] <- trends_func(station, variable)
    }
  }
}

```


#### Extract Regression Statistics into single table

```{r `regression stats`}
regression_statistics <- list()

for (station in names(trends_results)) {
      
  modelSums <- trends_results[[station]]
  
      reg.df <- data.frame(STATION_ID = station,
                           variable = vars[vars %in% names(yearlyAvgs_complete[[station]])],
               model_slope = rep(NA, length(yearlyAvgs_complete[[station]])),
               model_MAD = rep(NA, length(yearlyAvgs_complete[[station]])),
               model_pval = rep(NA, length(yearlyAvgs_complete[[station]])),
               model_intercept = rep(NA, length(yearlyAvgs_complete[[station]])))
      
  for (i in seq_along(yearlyAvgs_complete[[station]])) {
      
        p <- modelSums[[i]]
      
        reg.df$model_slope[i] <-  p$slope
        reg.df$model_MAD[i] <-  p$MAD
        reg.df$model_pval[i] <-  p$pvalue
        reg.df$model_intercept[i] <-  p$intercept
      }
      
      regression_statistics[[station]] <- reg.df
}
  

# ------- Create data frames from lists -------
regressions_df <- bind_rows(regression_statistics) 

# write_csv(regressions_df, "data/longTerm_Annual_Regression_Statistics.csv")

```

## Plotting

I initially created sample plot based on one station and variable and tried to incorporate the trend line using geom_abline. This worked in our reservoir plots. However, in this case the trend line did not appear. There was no error thrown, so I assume that for some reason the line was being plotted outside the bounds of the plot axes.

What may have happened:
geom_abline() expects an intercept and slope that apply globally across the x-axis, but if there's a mismatch between the scaling of the intercept and the x-axis (in this case, years starting at 2000), it may fail to display correctly.
By manually drawing the line with geom_segment(), I can specify the exact x and y coordinates, which resolves the issue.

#### Adjusted model intercept

Because the model intercept is based on x=0, and the x-axes are somewhere between 1980 and 200x, I created an adjusted model intercept based on the following formula:

`adjusted_intercept = model_intercept + model_slope * min(Year)`

This required first determining the minimum year for each variable at each station

```{r}
minYear <- list()

for (station in unique(regressions_df$STATION_ID)) {
  # Filter variables that exist in the current station
  station_vars <- vars[vars %in% names(yearlyAvgs_complete[[station]])]
  
  # Initialize the data frame for storing minimum years
  minYear.df <- data.frame(STATION_ID = station,
                           variable = station_vars,
                           minYear = rep(NA, length(station_vars)))
  
  # Loop through the variables and calculate the minimum year
  for (i in seq_along(station_vars)) {
    var <- station_vars[i]
    minYear.df$minYear[i] <- min(yearlyAvgs_complete[[station]][[var]]$Year, na.rm = TRUE)
  }
  
  # Store the data frame for this station
  minYear[[station]] <- minYear.df
}

minYear.df <- bind_rows(minYear)

```

Then using that data frame to calculate the adjusted intercept

```{r}
regressions_adjusted <- regressions_df %>%
  left_join(minYear.df, by = c("STATION_ID", "variable")) %>%
  mutate(adjusted_intercept = model_intercept + model_slope * minYear)

```



```{r, fig.width=3, fig.height=4}
library(ggplot2)
# use regressions_adjusted along with the monthly averages list
# monthly avg list structure: list[[station]]$variable  variable = df with Year, STATION_ID, MAY_OCT_AVG
# recall that, due to thow the output of the `mblm` function is structured, it cannot be passed to geom_smooth as a smoothing function. This results in having to 'manually' draw the trend lines.
# List to hold all plots 
all_plots_list <- list()

for (station in names(yearlyAvgs_complete)) {
  # Initiate list to store plots for this station
  station_plots <- list()
  
  for (current_var in names(yearlyAvgs_complete[[station]])) {
    df_avgs <- yearlyAvgs_complete[[station]][[current_var]]
    df_reg <- regressions_adjusted %>%
      filter(variable == current_var, STATION_ID == station)
    
    var_plot <- ggplot(df_avgs, aes(x = Year, y = MAY_OCT_AVG)) +
      geom_point() +
        # Use annotate() instead of geom_segment() for the trend line. ggplot expects each aesthetic  to be mapped to a column in the data, and geom_segment() - which I originally used - was receiving values (x, xend, y, and yend) that had only a single length (min and max). Also, aes() is designed to map aesthetics to a SINGLE dataframe, whereas this plot requires reference to two dfs to produce the trend lines.Using `annotate` eliminates the warning because it explicitly adds a line based on single values (rather than having ggplot2 try to match them to rows in df_avgs)
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = 
                 df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               # calculate the y-value at the endpoint by adding the total change in y (slope * difference_in_years) to the intercept at the starting point
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = current_var,
           subtitle = paste("Slope: ", round(df_reg$model_slope, 5),
                            "\np-value: ", round(df_reg$model_pval, 5)),
           x = "Year", y = NULL) +
      theme_minimal() +
      theme(plot.title = element_text(face = "bold"))
    
    station_plots[[current_var]] <- var_plot
  }

  all_plots_list[[station]] <- station_plots
}



save(all_plots_list, file = "all_plots_list.RData")

```

# -----------------------------------------------------------------------------

# Regressions for All Stations-Variables

After completing the above workflow, Paul requested that I generate a file with regression statistics (but without time series plots) for all stations-variables with at least 10 years of data based on May-Oct averages. In effect, do what was done above, but include all stations as opposed to the 33 stations of interest.

```{r}
# This file contains a column with the number of years of observations. 
metadata <- read_csv("Data/station_metadata.csv")

```

```{r}
# Function to subset data based on a minimum years of observations threshold. All station-variables in `data` have already been filtered to include only those with min 3 measurements per year (May-Oct).
subset_variable_data <- function(metadata_column, threshold = 10) {
  stations <- metadata %>%
    filter(!!sym(metadata_column) >= threshold) %>%
    pull(STATION_ID)
  
  return(data %>%
    filter(STATION_ID %in% stations))
}

# Create the subset_data list using the function
subset_data <- list(
  CHLa = subset_variable_data("nYears_CHLA"),
  PHOSPHORUS_TOTAL = subset_variable_data("nYears_totPhosphorus"),
  NITROGEN_TOTAL = subset_variable_data("nYears_totNitrogen"), 
  DIN = subset_variable_data("nYears_DIN"),
  SECCHI_DEPTH = subset_variable_data("nYears_Secchi"),   
  TSS = subset_variable_data("nYears_TSS")
)

# Derive May-Oct Means for each year by station-variable
stationAvgs <- list()

for (var in names(subset_data)) {
  df <- subset_data[[var]]
  
  # Group by Year and STATION_ID, and calculate the May-Oct average for the variable
  sy_avgs <- df %>%
    group_by(Year, STATION_ID) %>%
    summarise(MAY_OCT_AVG = mean(!!sym(var), na.rm = TRUE))
  
  stationAvgs[[var]] <- sy_avgs
}

```

## Run MBLM for each station and variable

Each station is stored as a list in `stationAvgs`. This list has a different structure from the one used above. Here, the list contains 6 dataframes, one for each variable. 

#### Custom function to derive trends

```{r, warning = FALSE}
trends_func2 <- function(variable) {
  
  modelSummaries <- list()  
  
  df <- stationAvgs[[variable]]
  
  for (station in unique(df$STATION_ID)) {
    # Filter for current station
    df_station <- df %>%
      filter(STATION_ID == station) %>%
      filter(!is.na(MAY_OCT_AVG)) %>%
      ungroup()
    
    # Run mblm model
    model <- mblm::mblm(MAY_OCT_AVG ~ Year, data = df_station)
    mod.sum <- mblm::summary.mblm(model)
    
    # Store results (directly, without creating the combined station-variable name)
    modelSummary <- list(
      station = station,
      slope = mod.sum$coefficients[2, 1],
      MAD = mod.sum$coefficients["Year", "MAD"],
      pvalue = mod.sum$coefficients["Year", 4],
      intercept = mod.sum$coefficients[1, 1]
    )
    
    modelSummaries[[station]] <- modelSummary
  }
  
  return(modelSummaries)  # Return the results
}

# ------- Run custom function in for loop -------

# Trends for all stations/variables with 10+ yr and 3+ obs/yr
trends_results2 <- list()

  for (variable in vars) {
  trends_results2[[variable]] <- trends_func2(variable)
  }


```


#### Extract Regression Statistics into single table

```{r `regression stats`}
regStats_allStations<- list()

for (variable in names(trends_results)) {
      
  modelSums <- trends_results[[variable]]
  
      reg.df <- data.frame(STATION_ID = names(modelSums),
                           variable = variable,
               model_slope = rep(NA, length(names(modelSums))),
               model_MAD = rep(NA, length(names(modelSums))),
               model_pval = rep(NA, length(names(modelSums))),
               model_intercept = rep(NA, length(names(modelSums))))
      
  for (i in seq_along(names(modelSums))) {
      
        p <- modelSums[[i]]
      
        reg.df$model_slope[i] <-  p$slope
        reg.df$model_MAD[i] <-  p$MAD
        reg.df$model_pval[i] <-  p$pvalue
        reg.df$model_intercept[i] <-  p$intercept
      }
      
      regStats_allStations[[variable]] <- reg.df
}
  

# ------- Create data frames from lists -------
regStats_allStations_df <- bind_rows(regStats_allStations) %>%
  arrange(STATION_ID)

# write_csv(regStats_allStations_df, "data/longTerm_Annual_Regression_Statistics_ALLSTATIONS.csv")

```

# ------------------------------------------------------------------------------


# Potential Publication Plots

"From among the station time series you plotted, I picked out 4 that I might want to use later as examples of trends.  I would like to get a version of these as high resolution jpegs.  For all of these, please include the y axis titles and make sure the y axis starts at zero.  Within each figure (station) I would like all the plots to have the same x axis range.  Lastly, please format the p values as follows: if p > 0.05, display as 0.xx, if p < 0.05 then display as 0.xxx, unless p <0.001, then display as p < 0.001.

The 4 stations are:

```
a. 4-ABWR010.55: a 2x3 figure with CHLa & TP at the top, TN and DIN in the middle, and TSS and Secchi at the bottom.
b. 9-PKC000.00 (as above)
c. 4-AROA192.94 (as above, but remove DIN and TSS outliers and re-run model).
d. 4-ASRE046.90 (as above but without DIN and TSS)."
```
All of these stations are included in the original 33 stations of interest, and so the original `regressions_adjusted` data frame can be used to create the plots. 

Station 4-AROA192.94 requires some additional data cleaning to remove outliers from the DIN and TSS variables.

Because there are only 4 stations, this can just be done manually to meet the individual specifications of each plot, rather than automating with a loop.

```{r}
# `cowplot` package for use in arranging the plots into single figures
library(cowplot)

```

Additional fields can be added to the data frame to facilitate the plotting process. This includes a field for defining p-val rendering and a field for y axis labeling.

```{r}
regressions_adjusted <- regressions_adjusted %>%
  mutate(pval_render = case_when(
    model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
    model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
    model_pval < 0.001 ~ "p < 0.001"
  ),
  y_label = case_when(
    variable == "CHLa" ~ "CHLa (\u03BCg L\u207B\u00B9)",
    variable == "PHOSPHORUS_TOTAL" ~ "TP (mg L\u207B\u00B9)",
    variable == "NITROGEN_TOTAL" ~ "TN (mg L\u207B\u00B9)",
    variable == "TSS" ~ "TSS (mg L\u207B\u00B9)",
    variable == "SECCHI_DEPTH" ~ "Secchi Depth (m)",
    variable == "DIN" ~ "DIN (mg L\u207B\u00B9)"
  ))

```


### a. 4-ABWR010.55

```{r, fig.width=6.5, fig.height=7}
station <- "4ABWR010.55"
data_a <- regressions_adjusted %>%
  filter(STATION_ID == "4ABWR010.55")

# List to hold the plots for this station
station_a <- list()

# Set a standard x-axis range across all plots (adjust as per data)
min_year <- 1980
max_year <- 2025  # Set the max year based on your data


# ! ---- Plot Generation ----- ! #
for (current_var in data_a$variable)  {
  df_avgs <- yearlyAvgs_complete[[station]][[current_var]]
  df_reg <- data_a %>%
    filter(variable == current_var, STATION_ID == station)
  
  # Extract the y_label for the current variable
  current_y_label <- df_reg$y_label[1]
  # max y value for dynamic range -- needed, since the ylim needs to be set at 0 on the low end
  max_y <- max(df_avgs$MAY_OCT_AVG, na.rm = TRUE) * 1.2
  
  # Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(df_avgs$Year) - 5  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis
  
  var_plot <- ggplot(df_avgs, aes(x = Year, y = MAY_OCT_AVG)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = df_reg$pval_render, size = 3.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x = "Year", y = current_y_label) +  # Use the extracted y_label
      scale_y_continuous(limits = c(0, max_y), labels = function(y) ifelse(y == 0, "0", format(y, digits = 1, nsmall = 0))) +
      # Set x-axis limits to be consistent
      scale_x_continuous(limits = c(min_year, max_year)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 9.5),        # Adjust axis title size if needed
        axis.text = element_text(size = 7.5, color = "black"),         # Adjust axis text size
        plot.subtitle = element_text(size = 9),     # Adjust subtitle size
        plot.margin = unit(c(.5, .5, .5, .5), "cm")
      )
  
  station_a[[current_var]] <- var_plot
}

  
  plot_4ABWR010.55 <- plot_grid(plotlist = station_a, ncol = 2,
                              vjust = 4)
  print(plot_4ABWR010.55)
  
  ggsave("plot_4ABWR010.55.jpg", width = 6.5, height = 7, units = "in", dpi = 1000)

```

### b. 9-PKC000.00

```{r, fig.width=6.5, fig.height=7}
station <- "9-PKC000.00"
data_b <- regressions_adjusted %>%
  filter(STATION_ID == station)

# List to hold the plots for this station
station_b <- list()

# Set a standard x-axis range across all plots (adjust as per data)
min_year <- 1980
max_year <- 2024  # Set the max year based on your data


# ! ---- Plot Generation ----- ! #
for (current_var in data_b$variable)  {
  df_avgs <- yearlyAvgs_complete[[station]][[current_var]]
  df_reg <- data_b %>%
    filter(variable == current_var, STATION_ID == station)
  
  # Extract the y_label for the current variable
  current_y_label <- df_reg$y_label[1]
  # max y value for dynamic range -- needed, since the ylim needs to be set at 0 on the low end
  max_y <- if (current_var == "CHLa") {
    15
  } else if (current_var == "PHOSPHORUS_TOTAL") {
    0.105 
  } else {
    max(df_avgs$MAY_OCT_AVG, na.rm = TRUE) * 1.2
  }
  
  # Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(df_avgs$Year) - 5  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis
  
  var_plot <- ggplot(df_avgs, aes(x = Year, y = MAY_OCT_AVG)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = df_reg$pval_render, size = 3.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x = "Year", y = current_y_label) +  # Use the extracted y_label
      scale_y_continuous(limits = c(0, max_y), labels = function(y) ifelse(y == 0, "0", format(y, digits = 1, nsmall = 0))) +
      # Set x-axis limits to be consistent
      scale_x_continuous(limits = c(min_year, max_year)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 9.5),        # Adjust axis title size if needed
        axis.text = element_text(size = 7.5, color = "black"),         # Adjust axis text size
        plot.subtitle = element_text(size = 9),     # Adjust subtitle size
        plot.margin = unit(c(.5, .5, .5, .5), "cm")
      )
  
  station_b[[current_var]] <- var_plot
}

  
  plot_9PKC000.00 <- plot_grid(plotlist = station_b, ncol = 2,
                              vjust = 4)
  print(plot_9PKC000.00)
  
  ggsave("plot_9PKC000.00.jpg", width = 6.5, height = 7, units = "in", dpi = 1000)

```

### c. 4-AROA192.94 

Outliers for DIN and TSS must be removed from yearly May-Oct averages, and MBLM models re-run. There is 1 outlier in each data set.

#### Data prep
```{r}
station <- "4AROA192.94"

stationData_DIN <- yearlyAvgs_complete[[station]][["DIN"]]
stationData_TSS <- yearlyAvgs_complete[[station]][["TSS"]]

boxplot(stationData_DIN$MAY_OCT_AVG, main = "DIN", ylab = "DIN (mg L-1)")
boxplot(stationData_TSS$MAY_OCT_AVG, main = "TSS", ylab = "TSS (mg L-1)")

# Remove outliers and return outlier-free data to the yearlyAvgs_complete list, for use in plotting loops
stationData_DIN <- stationData_DIN %>%
  filter(Year != 1983)
stationData_TSS <- stationData_TSS %>%
  filter(Year != 1989)

yearlyAvgs_complete[[station]][["DIN"]] <- stationData_DIN
yearlyAvgs_complete[[station]][["TSS"]] <- stationData_TSS

# Re-run MBLM regressions with outlier-free data (this uses the custom function )
  ## Replace the existing results in the trends_results list with outlier-free results
  trends_results[[station]]$DIN <- trends_func(station, "DIN")
  trends_results[[station]]$TSS <- trends_func(station, "TSS")

  ## Subset trend results for station 4-AROA192.94
  station_trends_results <- trends_results[[station]]


  ## Extract regression statistics into a single table
  reg.df <- data.frame(STATION_ID = station,
                       variable = vars[vars %in% names(station_trends_results)],
                       model_slope = rep(NA, length(names(station_trends_results))),
                       model_MAD = rep(NA, length(names(station_trends_results))),
                       model_pval = rep(NA, length(names(station_trends_results))),
                       model_intercept = rep(NA, length(names(station_trends_results)))
                     )
      
    for (i in seq_along(names(station_trends_results))) {
        
          p <- station_trends_results[[i]]
        
          reg.df$model_slope[i] <-  p$slope
          reg.df$model_MAD[i] <-  p$MAD
          reg.df$model_pval[i] <-  p$pvalue
          reg.df$model_intercept[i] <-  p$intercept
        }

  
# Calculate adjusted model intercept based on the minimum year per variable -- this is the same as above, and really only needs to be recalculated for DIN, for whom the min year has changed from 1983 to 1984.
## Change the min for DIN to 1984
station_minYears <- minYear.df %>%
  filter(STATION_ID == "4AROA192.94") %>%
  mutate(minYear = ifelse(variable == "DIN", 1984, minYear))

## Add adjsuted intercept, along with labeling and p-value rendering to the regression data frame
reg_adjustedIntcpt <- reg.df %>%
  left_join(station_minYears, by = c("STATION_ID", "variable")) %>%
  mutate(adjusted_intercept = model_intercept + model_slope * minYear) %>%
  mutate(pval_render = case_when(
    model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
    model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
    model_pval < 0.001 ~ "p < 0.001"
  ),
  y_label = case_when(
    variable == "CHLa" ~ "CHLa (\u03BCg L\u207B\u00B9)",
    variable == "PHOSPHORUS_TOTAL" ~ "TP (mg L\u207B\u00B9)",
    variable == "NITROGEN_TOTAL" ~ "TN (mg L\u207B\u00B9)",
    variable == "TSS" ~ "TSS (mg L\u207B\u00B9)",
    variable == "SECCHI_DEPTH" ~ "Secchi Depth (m)",
    variable == "DIN" ~ "DIN (mg L\u207B\u00B9)"
  ))

```

#### Plot generation

```{r, fig.width=6.5, fig.height=7}
# List to hold the plots for this station
station_c <- list()

# Set a standard x-axis range across all plots (adjust as per data)
min_year <- 1983
max_year <- 2024  # Set the max year based on your data


# ! ---- Plot Generation ----- ! #
for (current_var in reg_adjustedIntcpt$variable)  {
  df_avgs <- yearlyAvgs_complete[[station]][[current_var]]
  df_reg <- reg_adjustedIntcpt %>%
    filter(variable == current_var)
  
  # Extract the y_label for the current variable
  current_y_label <- df_reg$y_label[1]
  # max y value for dynamic range -- needed, since the ylim needs to be set at 0 on the low end
   max_y <- if (current_var == "CHLa") {
      30
    } else {
      max(df_avgs$MAY_OCT_AVG, na.rm = TRUE) * 1.2
    }
    
   print(paste("Current variable:", current_var, "Max Y:", max_y))
   
  # Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(df_avgs$Year) - 5  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis
  
  var_plot <- ggplot(df_avgs, aes(x = Year, y = MAY_OCT_AVG)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = df_reg$pval_render, size = 3.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x = "Year", y = current_y_label) +  # Use the extracted y_label
      scale_y_continuous(limits = c(0, max_y), labels = function(y) ifelse(y == 0, "0", format(y, digits = 1, nsmall = 0))) +
      # Set x-axis limits to be consistent
      scale_x_continuous(limits = c(min_year, max_year)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 9.5),        # Adjust axis title size if needed
        axis.text = element_text(size = 7.5, color = "black"),         # Adjust axis text size
        plot.subtitle = element_text(size = 9),     # Adjust subtitle size
        plot.margin = unit(c(.5, .5, .5, .5), "cm")
      )
  
  station_c[[current_var]] <- var_plot
}

  
  plot_4AROA192.94 <- plot_grid(plotlist = station_c, ncol = 2,
                              vjust = 4)
  print(plot_4AROA192.94)
  
  ggsave("plot_4AROA192.94.jpg", width = 6.5, height = 7, units = "in", dpi = 1000)

```

### d. 4-ASRE046.90

This figure should include only 4 plots, with DIN and TSS excluded.

```{r, fig.width=6.5, fig.height=4.667}
station <- "4ASRE046.90"
data_d <- regressions_adjusted %>%
  filter(STATION_ID == station) %>%
  filter(variable != "DIN" & variable != "TSS")

# List to hold the plots for this station
station_d <- list()

# Set a standard x-axis range across all plots (adjust as per data)
min_year <- 1980
max_year <- 2024  # Set the max year based on your data


# ! ---- Plot Generation ----- ! #
for (current_var in data_d$variable)  {
  df_avgs <- yearlyAvgs_complete[[station]][[current_var]]
  df_reg <- data_d %>%
    filter(variable == current_var, STATION_ID == station)
  
  # Extract the y_label for the current variable
  current_y_label <- df_reg$y_label[1]
  # max y value for dynamic range -- needed, since the ylim needs to be set at 0 on the low end
  max_y <- max(df_avgs$MAY_OCT_AVG, na.rm = TRUE) * 1.07

  
  # Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- if (current_var == "PHOSPHORUS_TOTAL") {
    max(df_avgs$Year) - 18.5
    } else {
      max(df_avgs$Year) - 10
    }
      
  pval_y_pos <- if (current_var == "PHOSPHORUS_TOTAL") {
    .023
    } else if (current_var == "NITROGEN_TOTAL") {
      .345
    } else if (current_var == "CHLa") {
      2.5
    } else {
      max_y * 0.95
    }         
  
  var_plot <- ggplot(df_avgs, aes(x = Year, y = MAY_OCT_AVG)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = df_reg$pval_render, size = 3.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x = "Year", y = current_y_label) +  # Use the extracted y_label
      scale_y_continuous(limits = c(0, max_y), labels = function(y) ifelse(y == 0, "0", format(y, digits = 1, nsmall = 0))) +
      # Set x-axis limits to be consistent
      scale_x_continuous(limits = c(min_year, max_year)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 9.5),        # Adjust axis title size if needed
        axis.text = element_text(size = 7.5, color = "black"),         # Adjust axis text size
        plot.subtitle = element_text(size = 9),     # Adjust subtitle size
        plot.margin = unit(c(.5, .5, .5, .5), "cm")
      )
  
  station_d[[current_var]] <- var_plot
}

  
  plot_4ASRE046.90 <- plot_grid(plotlist = station_d, ncol = 2,
                              vjust = 4)
  print(plot_4ASRE046.90)
  
  ggsave("plot_4ASRE046.90.jpg", width = 6.5, height = 4.667, units = "in", dpi = 1000)
  
  
  d <- bind_rows(yearlyAvgs_complete[[station]], .id = "variable")
  
  d %>%
    group_by(variable) %>%
    summarise(range = range(MAY_OCT_AVG, na.rm = TRUE))

```
