---
title: "Reservoir Water Clatity: Population Level Analysis"
output: html_document
date: "2024-10-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

```{r, message = FALSE}
# Load the data
data <- read_csv("data/processed_dataset.csv")

  ## Remove 2024 obs since they are incomplete
  data <- data %>%
    filter(Year != 2024)
    
# This file contains a column with the number of years of observations. 
metadata <- read_csv("Data/station_metadata.csv")

```

# Subset Stations of Interest

 "For this, we will be working with the subset of stations that have at least 5 years of data (inclusive of those used in the trend analysis).  From the station metadata, I estimate there are ~175 stations that meet this minimum for Secchi, CHLa, TN and TP (~40 for TSS)."  (*Still min 3 Nobs, presumably*??)
 
```{r}
# Function to subset data based on a minimum years of observations threshold. All station-variables in `data` have already been filtered to include only those with min 3 measurements per year (May-Oct). 
# In this case, 5 years of observations is the threshold.
subset_variable_data <- function(metadata_column, threshold = 5) {
  stations <- metadata %>%
    filter(!!sym(metadata_column) >= threshold) %>%
    pull(STATION_ID)
  
  return(data %>%
    filter(STATION_ID %in% stations))
}

# Create the subset_data list using the function
subset_data <- list(
  SECCHI_DEPTH = subset_variable_data("nYears_Secchi"),   
  CHLa = subset_variable_data("nYears_CHLA"),           
  NITROGEN_TOTAL = subset_variable_data("nYears_totNitrogen"), 
  PHOSPHORUS_TOTAL = subset_variable_data("nYears_totPhosphorus"), 
  TSS = subset_variable_data("nYears_TSS"),
  DIN = subset_variable_data("nYears_DIN")
)

# Remove outliers
## DIN outlier from station 4AROA192.94
din_data <- subset_data[["DIN"]]
index <- which.max(din_data$DIN[din_data$STATION_ID == "4AROA192.94"])
din_data$DIN[din_data$STATION_ID == "4AROA192.94"][index] <- NA
subset_data[["DIN"]] <- din_data

## CHLa outliers
### station 2-XDD000.40 (value = 1148) and station 5ASRN000.66 (value = 11592)
chla_data <- subset_data[["CHLa"]]
index1 <- which.max(chla_data$CHLa[chla_data$STATION_ID == "2-XDD000.40"])
index2 <- which.max(chla_data$CHLa[chla_data$STATION_ID == "5ASRN000.66"])
chla_data$CHLa[chla_data$STATION_ID == "2-XDD000.40"][index1] <- NA
chla_data$CHLa[chla_data$STATION_ID == "5ASRN000.66"][index2] <- NA
subset_data[["CHLa"]] <- chla_data

```

# Summary Statistics
 
"I would like to get a file containing the long-term min, mean, median and max for each of the 6 variables (CHLa, TSS, etc.) by station for all the stations that meet the minimum of 5 years criterion."

```{r}
# Function to calculate summary statistics for each variable by station
calculate_summary_stats <- function(data, variable) {
  return(data %>%
    group_by(STATION_ID) %>%
    summarise(
      min = min(!!sym(variable), na.rm = TRUE),
      mean = mean(!!sym(variable), na.rm = TRUE),
      median = median(!!sym(variable), na.rm = TRUE),
      max = max(!!sym(variable), na.rm = TRUE)
    ))
}


summary_list <- list() # List to hold summary stats for each variable

# Use above function in for loop, storing results in list
for (var in names(subset_data)) {
  df <- subset_data[[var]]
  summary_stats <- calculate_summary_stats(df, var)
  
  summary_list[[var]] <- summary_stats
}

# Bind into single df
summary_df <- bind_rows(summary_list, .id = "variable") %>%
  relocate(variable, .after = "STATION_ID")

## "organize these data so that there is a single column of Station IDs followed by 4 columns each (min, mean, median, max) for CHLa, TN, TP, etc.
## That way there is a single row per station containing all the statistics for the 6 variables"
summary_df_wide <- summary_df %>%
  pivot_wider(
    names_from = variable,
    values_from = c(min, mean, median, max),
    names_glue = "{variable}_{.value}"
  ) %>%
  select(
    STATION_ID,
    matches("^CHLa_"), 
    matches("^SECCHI_DEPTH_"), 
    matches("^NITROGEN_TOTAL_"), 
    matches("^PHOSPHORUS_TOTAL_"), 
    matches("^TSS_"), 
    matches("^DIN_")
  )

summary_df_wide

write_csv(summary_df_wide, "data/5y_summaryStats_wide.csv")

```

# Box-Whisker Plots

 Next, I would like to see box-whisker plots for each variable with the data partitioned into 5-y increments.  For convenience, I would suggest making the last box a 4-y increment (2020-23) and then go backwards in 5-y increments (2000-04, 2005-09, 2010-14, 2015-19, etc.).  As we go back past 2000, the data become more sparse.  *It would be helpful to indicate on the plot the number of observations used to make each box.  We will likely adopt a cut-off below which we will not include that interval/variable, but I don't know offhand what that should be.*


### Partition Data - 5y Increments

Working backwards from 2023, with first increment (2019-2023) being 4 years.

```{r}
# --- Custom function --- #
# This function creates intervals based on the min year of the variable, while also handling the 2019-2023 interval
## Vectorize is needed because the function was written inititally to only work on single values
## `interval_start <- floor((year - min_year) / 5) * 5 + min_year` -- > This line of code calculates the starting year of a 5-year interval based on the given year and the earliest year in the dataset (min_year):

   ## year - min_year: Determines how many years have passed since the minimum year in the dataset.
   ## / 5 and floor(): Divides the difference by 5 and rounds down to group the year into a 5-year block.
   ##  * 5: Converts the block number back to the appropriate number of years (in multiples of 5).
   ##  + min_year: Shifts the result back to the original time scale by adding the minimum year.

create_intervals <- function(year) {
  case_when(
    year >= 2020 ~ "2020-2023",
    year >= 2015 ~ "2015-2019",
    year >= 2010 ~ "2010-2014",
    year >= 2005 ~ "2005-2009",
    year >= 2000 ~ "2000-2004",
    year >= 1995 ~ "1995-1999",
    year >= 1990 ~ "1990-1994",
    year >= 1985 ~ "1985-1989",
    year >= 1980 ~ "1980-1984",
    year >= 1975 ~ "1975-1979",
    year >= 1972 ~ "1972-1974",
    TRUE ~ NA
  )
}



# Use function to assign each May-Oct avg to an interval, by variable
boxplot_data <- list()
for (var in names(subset_data)) {
  x <- subset_data[[var]] %>%
    filter(!is.na(!!sym(var))) 
  min_year <- min(x$Year)
  
  x_w_intervals <- x %>%
    mutate(interval = create_intervals(Year))
  
  boxplot_data[[var]] <- x_w_intervals %>%
        group_by(STATION_ID, Year) %>%
        summarize(yearly_mean = mean(!!sym(var), na.rm = TRUE),
                  interval = first(interval))

}

```

How many observations per interval, per variable? This informatio needs to be incorporated into the plots.

```{r}
# List of Nobs per interval for incorporation into plots
obs_counts <- list()
for (var in names(boxplot_data)) {
  x <- boxplot_data[[var]]
  x_counts <- x %>%
    group_by(interval) %>%
    summarise(n = n())
  
  obs_counts[[var]] <- x_counts
}

```

## Interval statistics

Generate a file of the statistics associated with the box-whisker results (N Obs and median values for each span of years shown in the plots).

```{r}
summarize_intervals <- function(x, y) {
  x %>%
    left_join(y, by = "interval") %>%
    group_by(interval) %>%
    summarise(
      n = first(n),
      median = median(yearly_mean, na.rm = TRUE)
    )
}

interval_summaries <- list()
for (var in names(boxplot_data)) {
  interval_summaries[[var]] <- summarize_intervals(boxplot_data[[var]], obs_counts[[var]])
}

interval_summary.df <- bind_rows(interval_summaries, .id = "variable") %>%
  relocate(variable, .after = "interval")

write_csv(interval_summary.df, "data/population_boxplot_interval_summaries.csv")

```

## Create Box-Whisker Plots

```{r}
library(ggplot2)
 
# ------------ SECCHI DEPTH  -------------------------
  plot_secchi <- ggplot(boxplot_data[["SECCHI_DEPTH"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
    geom_jitter(width = 0.1, alpha = 0.2, size = 0.7) + 
    geom_text(data = obs_counts[["SECCHI_DEPTH"]], aes(x = interval, y = max(boxplot_data[["SECCHI_DEPTH"]]$yearly_mean, na.rm = TRUE), 
                                   label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  # Add observation counts
    labs(
      title = "Secchi Depth",
      subtitle = "No outliers excluded for visualization.",
      x = NULL,
      y = "Depth (m)"
    ) +
    theme_minimal() +
    theme(legend.position = "none",
          plot.subtitle = element_text(size = 10),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 10))  
  
  print(plot_secchi)
  ggsave("figures/secchi_boxplot.jpg", plot_secchi, width = 7, height = 5, units = "in", dpi = 600)
  
# ------------- Chlorophyll-A  -------------------------
  ## Filter out CHLa values above 300 to minimize y axis distortion
  boxplot_data[["CHLa"]] <- boxplot_data[["CHLa"]] %>%
    filter(yearly_mean <= 50)
  
  plot_chla <- ggplot(boxplot_data[["CHLa"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
    geom_jitter(width = 0.1, alpha = 0.2, size = 0.7) +  
    geom_text(data = obs_counts[["CHLa"]], aes(x = interval, y = max(boxplot_data[["CHLa"]]$yearly_mean, na.rm = TRUE), 
                                   label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  # Add observation counts
    labs(
      title = "Chlorophyll-A",
      subtitle = "CHLa values above 50 (n = 63) excluded for visualization purposes.",
      x = NULL,
      y = "CHLa (\u03BCg L\u207B\u00B9)"
    ) +
    theme_minimal() +
    theme(legend.position = "none",
          plot.subtitle = element_text(size = 10),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 10))    
  
  print(plot_chla)
  ggsave("figures/chla_boxplot.jpg", plot_chla, width = 7, height = 5, units = "in", dpi = 600)
  
    # ------------ CHLa log base 10 transformed -------------------------
        trnsfrm.data <- boxplot_data[["CHLa"]] %>%
          mutate(log10 = log10(yearly_mean))
  
          # Filter outliers above 2 to minimize y axis distortion
          trnsfrm.data <- trnsfrm.data %>%
            filter(log10 <= 2)
  
        plot_chlalog10 <- ggplot(trnsfrm.data, aes(x = interval, y = log10)) +
          geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
          geom_jitter(width = 0.1, alpha = 0.2, size = 0.7) +  
          geom_text(data = obs_counts[["CHLa"]], aes(x = interval, y = max(trnsfrm.data$log10, na.rm = TRUE), 
                                         label = paste("n =", n)), 
                  vjust = -0.5, size = 3, color = "black") +  # Add observation counts
          labs(
            title = "Chlorophyll-A (log\u2081\u2080 transformed)",
            subtitle = "log\u2081\u2080(CHLa) values above 2 (n = 11) excluded for visualization purposes.",
            x = NULL,
            y = "log\u2081\u2080(CHLa) (\u03BCg L\u207B\u00B9)"
          ) +
          theme_minimal() +
          theme(legend.position = "none",
                plot.subtitle = element_text(size = 10),
                axis.text.x = element_text(angle = 45, hjust = 1, size = 10))   
        
        print(plot_chlalog10)
        ggsave("figures/log10_chla_boxplot.jpg", plot_chlalog10, width = 7, height = 5, units = "in", dpi = 600)
    

# ------------- Total Nitrogen -------------------------
  ## Filter out TN values above 2 to minimize y axis distortion
  boxplot_data[["NITROGEN_TOTAL"]] <- boxplot_data[["NITROGEN_TOTAL"]] %>%
    filter(yearly_mean <= 2)
  
  plot_tn <- ggplot(boxplot_data[["NITROGEN_TOTAL"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.2, size = 0.7) +
    geom_text(data = obs_counts[["NITROGEN_TOTAL"]], 
              aes(x = interval, y = max(boxplot_data[["NITROGEN_TOTAL"]]$yearly_mean, na.rm = TRUE), 
                                                       label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  
    labs(
      title = "Total Nitrogen",
      subtitle = "TN values above 2 (n = 28) excluded for visualization purposes.",
      x = NULL,
      y = "TN (mg N L\u207B\u00B9)"
  ) +
    theme_minimal() +
    theme(legend.position = "none",
        plot.subtitle = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))   
  
  print(plot_tn)
  ggsave("figures/tn_boxplot.jpg", plot_tn, width = 7, height = 5, units = "in", dpi = 600)

# ------------- Total Phosphorus -------------------------
  ## Filter out TP values above .3 to minimize y axis distortion
  boxplot_data[["PHOSPHORUS_TOTAL"]] <- boxplot_data[["PHOSPHORUS_TOTAL"]] %>%
    filter(yearly_mean <= .3)
  
  plot_tp <- ggplot(boxplot_data[["PHOSPHORUS_TOTAL"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
    geom_jitter(width = 0.15, alpha = 0.2, size = 0.7) +
    geom_text(data = obs_counts[["PHOSPHORUS_TOTAL"]], 
              aes(x = interval, y = max(boxplot_data[["PHOSPHORUS_TOTAL"]]$yearly_mean, na.rm = TRUE), 
                                                       label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  
    labs(
      title = "Total Phosphorus",
      subtitle = "TP values above .3 (n = 21) excluded for visualization purposes.",
      x = NULL,
      y = "TP (mg N L\u207B\u00B9)"
  ) +
    theme_minimal() +
    theme(legend.position = "none",
        plot.subtitle = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))    
  
  print(plot_tp)
  ggsave("figures/tp_boxplot.jpg", plot_tp, width = 7, height = 5, units = "in", dpi = 600)
  
# ------------- Total Suspended Solids-------------------------
  ## Filter out TSS values above 50 (n = 34) to minimize y axis distortion
  boxplot_data[["TSS"]] <- boxplot_data[["TSS"]] %>%
    filter(yearly_mean <= 50)
  
  plot_tss <- ggplot(boxplot_data[["TSS"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
    geom_jitter(width = 0.15, alpha = 0.2, size = 0.7) +
    geom_text(data = obs_counts[["TSS"]], 
              aes(x = interval, y = max(boxplot_data[["TSS"]]$yearly_mean, na.rm = TRUE), 
                                                       label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  
    labs(
      title = "Total Suspened Solids",
      subtitle = "TSS values above 50 (n = 6) excluded for visualization purposes.",
      x = NULL,
      y = "TSS (mg N L\u207B\u00B9)"
  ) +
    theme_minimal() +
    theme(legend.position = "none",
        plot.subtitle = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))    
  
  print(plot_tss)
  ggsave("figures/tss_boxplot.jpg", plot_tss, width = 7, height = 5, units = "in", dpi = 600)

```

# MBLM Trend Estimates

"for the dataset that contains all the stations which have a min of 5 years of data, I would like to fit mblms to estimate the trend (slope) for CHLa, TN, TP, Sechi, TSS and DIN.  This is the same as what we did previously for individual stations, but now we are fitting a model for all the stations combined including those stations that met the 5 year min but not the 10 or 15 year min.  It would be helpful to see a plot of the data with the model."


## Derive May-Oct averages

Yearly means for each variable at each station where 5y or more of data (min 3 obs/yr). 

```{r, message=FALSE}
vars <- names(subset_data)

# List to store station averages for each variable
stationAvgs <- list()

# Loop by variable, as `subset_data` is structured by variable
for (var in vars) {
  
  # List to hold averages for the current variable
  station_avgs_current_var <- list()
  
  var_df <- subset_data[[var]]
  stations <- unique(df$STATION_ID)
  
  for (station in stations) {
    # Filter for current station and include only non-NA values of current var
    station_df <- var_df %>%
      filter(STATION_ID == station) %>%
      filter(!is.na(!!sym(var))) # dynamically reference the variable
             
    # Group observations by year and filter for only years with 3 or more obs
    # This is redundant, as the processed dataset generated in pt1 file already filters for min 3 obs per station-year. But no harm in being extra cautious.
      years_with_3_obs <- station_df %>%
        group_by(Year) %>%
        summarise(n = sum(!is.na(!!sym(var)))) %>%
        filter(n >= 3)
      
        # Calculate May-Oct average for the current station and variable, filtering to only include years with 3+ obs
        df_avg <- station_df %>%
          filter(Year %in% years_with_3_obs$Year) %>% # Only include years with >= 3 obs
          group_by(Year, STATION_ID) %>%
          summarise(MAY_OCT_AVG = mean(!!sym(var), na.rm = TRUE))

        # Add the May-Oct average for variable to the station_avgs list
        station_avgs_current_var[[station]] <- df_avg
  }
  
  # Only add station to the final list if it has valid averages for at least one variable
  if (length(station_avgs_current_var) > 0) {
    stationAvgs[[var]] <- station_avgs_current_var
  }
}

# Create data frame to save and share with Paul
avg_df_list <- list()

for (var in vars) {
  df <- bind_rows(stationAvgs[[var]])
  avg_df_list[[var]] <- df
}

all_vars_avg_df <- bind_rows(avg_df_list, .id = "variable")

write_csv(all_vars_avg_df, "data/population_level_data/population_avg_data.csv")

```

## MBLM Model Fitting

```{r}
# Variable order within station average list
# [1] "SECCHI_DEPTH"     "CHLa"             "NITROGEN_TOTAL"  
# [4] "PHOSPHORUS_TOTAL" "TSS"              "DIN"

# Create function to loop through each variable and fit mblm
mblm_func <- function(variable) {
  
  averages_list <- stationAvgs[[variable]]
  # Bind into single data frame for use in mblm
  df <- bind_rows(averages_list)
  
    # Run mblm model
    model <- mblm::mblm(MAY_OCT_AVG ~ Year, data = df)
    mod.sum <- mblm::summary.mblm(model)
    
    # Store results (directly, without creating the combined station-variable name)
    modelSummary <- list(
      slope = mod.sum$coefficients[2, 1],
      MAD = mod.sum$coefficients["Year", "MAD"],
      pvalue = mod.sum$coefficients["Year", 4],
      intercept = mod.sum$coefficients[1, 1]
    )
    
    modelSummaries <- modelSummary
  
  return(modelSummaries)
}


# ------- Run custom function in for loop -------

# Store population-level trend estimates for each variable in list
trends_results <- list()

  for (var in vars) {
  trends_results[[var]] <- mblm_func(var)
  }


```


## Extract Regression Statistics

```{r `regression stats`}
regStats<- list()

for (variable in names(trends_results)) {
      
  modelSums <- trends_results[[variable]]
  
      reg.df <- data.frame(variable = variable,
               model_slope = NA,
               model_MAD = NA,
               model_pval = NA,
               model_intercept = NA)
      
        p <- modelSums
      
        reg.df$model_slope <-  p$slope
        reg.df$model_MAD <-  p$MAD
        reg.df$model_pval <-  p$pvalue
        reg.df$model_intercept <-  p$intercept
      
      regStats[[variable]] <- reg.df
}
  

# ------- Create data frames from lists -------
regStats_allVars <- bind_rows(regStats) 
```

### Visualize Models

#### Adjusted model intercept

Because the model intercept is based on x=0, and the x-axes are somewhere between 197X and 200x, I create an adjusted model intercept based on the following formula:

`adjusted_intercept = model_intercept + model_slope * min(Year)`

This requires first determining the minimum year for each variable at each station

```{r}
minYear <- list()

for (var in vars) {
  
  df <- bind_rows(stationAvgs[[var]])
  
  # Initialize the data frame for storing minimum years
  minYear.df <- data.frame(variable = var,
                           minYear = min(df$Year))
  
  # Store the data frame for this station
  minYear[[var]] <- minYear.df
}

minYear.df <- bind_rows(minYear)

```

Then using that data frame to calculate the adjusted intercept

```{r}
regressions_adjusted <- regStats_allVars %>%
  left_join(minYear.df, by = c("variable")) %>%
  mutate(adjusted_intercept = model_intercept + model_slope * minYear)

write_csv(regressions_adjusted, "data/population_level_data/population_trend_estimates.csv")

```

#### Plotting

```{r, echo=FALSE}
library(ggplot2)
# use regressions_adjusted along with the station averages list
# recall that, due to thow the output of the `mblm` function is structured, it cannot be passed to geom_smooth as a smoothing function. This results in having to 'manually' draw the trend lines.

# List to hold all plots 
all_plots_list <- list()
  
  for (current_var in names(stationAvgs)) {
    
    df_avgs <- stationAvgs[[current_var]] %>%
      bind_rows()
    
    df_reg <- regressions_adjusted %>%
      filter(variable == current_var)
    
    var_plot <- ggplot(df_avgs, aes(x = Year, y = MAY_OCT_AVG)) +
      geom_point(alpha = .3, position = position_jitter(width = 0.3)) +
        # Use annotate() instead of geom_segment() for the trend line. ggplot expects each aesthetic  to be mapped to a column in the data, and geom_segment() - which I originally used - was receiving values (x, xend, y, and yend) that had only a single length (min and max). Also, aes() is designed to map aesthetics to a SINGLE dataframe, whereas this plot requires reference to two dfs to produce the trend lines.Using `annotate` eliminates the warning because it explicitly adds a line based on single values (rather than having ggplot2 try to match them to rows in df_avgs)
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = 
                 df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               # calculate the y-value at the endpoint by adding the total change in y (slope * difference_in_years) to the intercept at the starting point
               color = "red", linewidth = .6, alpha = .8) +
      labs(title = current_var,
           subtitle = paste("Slope: ", round(df_reg$model_slope, 5),
                            "\np-value: ", round(df_reg$model_pval, 5)),
           x = "Year", y = NULL) +
      theme_minimal() +
      theme(plot.title = element_text(face = "bold"))

  all_plots_list[[current_var]] <- var_plot
}


# The plots are generated, now patch them together.
library(cowplot)

  # Use cowplot::plot_grid to combine plots 
  combined_plot <- plot_grid(plotlist = all_plots_list, ncol = 2)
  print(combined_plot)
  
  ggsave("figures/population_trend_plots.jpg", combined_plot, height = 10, units = "in", dpi = 600)

```