---
title: "Box plot/MBLM Trend Overlay"
output: html_document
date: "2024-10-09"
---

[This file is a copied and modified version of pt3_population_level_analysis
I wanted to keep that file intact, as the visualization request involves removing certain data points from the original dataset.]

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

"There is another thing I would like to try with the population data, but I'm not sure it will work.
I was wondering if you could take the model results from the plots you just sent, and overlay the trend lines on the box plots you made previously.  The aggregated data (box plots at 5-y intervals) are more useful for visualizing the data, but it would be helpful to also show how the model trend looks.
In doing this, I would like to re-run the mblm for Secchi and CHLa excluding the early data (1980-84), so the trend is based on data collected since 1995, and the model line added to the box plot starts with the 1995 group.
For TSS, I would like to fit the mblm to data spanning 1980 to 2005, which is where we have the most data (the model trend line would span the 5 boxes corresponding to this period).
For TN and TP, I would like to use all the data (i.e., the model you have already created)."

```{r, message = FALSE}
# Load the data
data <- read_csv("data/processed_dataset.csv")

  ## Remove 2024 obs since they are incomplete
  data <- data %>%
    filter(Year != 2024)
    
# This file contains a column with the number of years of observations. 
metadata <- read_csv("Data/station_metadata.csv")

```

# Subset Stations of Interest

 "For this, we will be working with the subset of stations that have at least 5 years of data (inclusive of those used in the trend analysis).  From the station metadata, I estimate there are ~175 stations that meet this minimum for Secchi, CHLa, TN and TP (~40 for TSS)."  (*Still min 3 Nobs, presumably*??)
 
```{r `subset data`}
# Function to subset data based on a minimum years of observations threshold. All station-variables in `data` have already been filtered to include only those with min 3 measurements per year (May-Oct). 
# In this case, 5 years of observations is the threshold.
subset_variable_data <- function(metadata_column, threshold = 5) {
  stations <- metadata %>%
    filter(!!sym(metadata_column) >= threshold) %>%
    pull(STATION_ID)
  
  return(data %>%
    filter(STATION_ID %in% stations))
}

# Create the subset_data list using the function
subset_data <- list(
  SECCHI_DEPTH = subset_variable_data("nYears_Secchi"),   
  CHLa = subset_variable_data("nYears_CHLA"),           
  NITROGEN_TOTAL = subset_variable_data("nYears_totNitrogen"), 
  PHOSPHORUS_TOTAL = subset_variable_data("nYears_totPhosphorus"), 
  TSS = subset_variable_data("nYears_TSS"),
  DIN = subset_variable_data("nYears_DIN")
)
```

### Outlier Removal
```{r}
## DIN outlier from station 4AROA192.94
din_data <- subset_data[["DIN"]]
index <- which.max(din_data$DIN[din_data$STATION_ID == "4AROA192.94"])
din_data$DIN[din_data$STATION_ID == "4AROA192.94"][index] <- NA
subset_data[["DIN"]] <- din_data

## CHLa outliers
### station 2-XDD000.40 (value = 1148) and station 5ASRN000.66 (value = 11592)
chla_data <- subset_data[["CHLa"]]
index1 <- which.max(chla_data$CHLa[chla_data$STATION_ID == "2-XDD000.40"])
index2 <- which.max(chla_data$CHLa[chla_data$STATION_ID == "5ASRN000.66"])
chla_data$CHLa[chla_data$STATION_ID == "2-XDD000.40"][index1] <- NA
chla_data$CHLa[chla_data$STATION_ID == "5ASRN000.66"][index2] <- NA
subset_data[["CHLa"]] <- chla_data



## Remove data from pre-1995 for CHLa and Secchi
for (var in c("CHLa", "SECCHI_DEPTH")) {
  subset_data[[var]] <- subset_data[[var]] %>%
    filter(Year >= 1995)
}

## Filter to only include 1980-2005 for TSS
subset_data[["TSS"]] <- subset_data[["TSS"]] %>%
  filter(Year >= 1980, Year <= 2005)

```


# Box-Whisker Plots

 Next, I would like to see box-whisker plots for each variable with the data partitioned into 5-y increments.  For convenience, I would suggest making the last box a 4-y increment (2020-23) and then go backwards in 5-y increments (2000-04, 2005-09, 2010-14, 2015-19, etc.).  As we go back past 2000, the data become more sparse.  *It would be helpful to indicate on the plot the number of observations used to make each box.  We will likely adopt a cut-off below which we will not include that interval/variable, but I don't know offhand what that should be.*


### Partition Data - 5y Increments

Working backwards from 2023, with first increment (2019-2023) being 4 years.

```{r}
create_intervals <- function(year) {
  case_when(
    year >= 2020 ~ "2020-2023",
    year >= 2015 ~ "2015-2019",
    year >= 2010 ~ "2010-2014",
    year >= 2005 ~ "2005-2009",
    year >= 2000 ~ "2000-2004",
    year >= 1995 ~ "1995-1999",
    year >= 1990 ~ "1990-1994",
    year >= 1985 ~ "1985-1989",
    year >= 1980 ~ "1980-1984",
    year >= 1975 ~ "1975-1979",
    year >= 1972 ~ "1972-1974",
    TRUE ~ NA
  )
}



# Use function to assign each May-Oct avg to an interval, by variable
boxplot_data <- list()
for (var in names(subset_data)) {
  x <- subset_data[[var]] %>%
    filter(!is.na(!!sym(var))) 
  min_year <- min(x$Year)
  
  x_w_intervals <- x %>%
    mutate(interval = create_intervals(Year))
  
  boxplot_data[[var]] <- x_w_intervals %>%
        group_by(STATION_ID, Year) %>%
        summarize(yearly_mean = mean(!!sym(var), na.rm = TRUE),
                  interval = first(interval))

}

```

How many observations per interval, per variable? This information needs to be incorporated into the plots.

```{r}
# List of Nobs per interval for incorporation into plots
obs_counts <- list()
for (var in names(boxplot_data)) {
  x <- boxplot_data[[var]]
  x_counts <- x %>%
    group_by(interval, interval_midpoint) %>%
    summarise(n = n())
  
  obs_counts[[var]] <- x_counts
}

```


## Create Box-Whisker Plots

```{r}
library(ggplot2)
 
# ------------ SECCHI DEPTH  -------------------------
  plot_secchi <- ggplot(boxplot_data[["SECCHI_DEPTH"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
    geom_jitter(width = 0.1, alpha = 0.2, size = 0.7) + 
    geom_text(data = obs_counts[["SECCHI_DEPTH"]], aes(x = interval, y = max(boxplot_data[["SECCHI_DEPTH"]]$yearly_mean, na.rm = TRUE), 
                                   label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  # Add observation counts
    labs(
      title = "Secchi Depth",
      subtitle = "No outliers excluded for visualization.",
      x = NULL,
      y = "Depth (m)"
    ) +
    theme_minimal() +
    theme(legend.position = "none",
          plot.subtitle = element_text(size = 10),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 10))  
  
  print(plot_secchi)
  
# ------------- Chlorophyll-A  -------------------------
  ## Filter out CHLa values above 300 to minimize y axis distortion
  boxplot_data[["CHLa"]] <- boxplot_data[["CHLa"]] %>%
    filter(yearly_mean <= 50)
  
  plot_chla <- ggplot(boxplot_data[["CHLa"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
    geom_jitter(width = 0.1, alpha = 0.2, size = 0.7) +  
    geom_text(data = obs_counts[["CHLa"]], aes(x = interval, y = max(boxplot_data[["CHLa"]]$yearly_mean, na.rm = TRUE), 
                                   label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  # Add observation counts
    labs(
      title = "Chlorophyll-A",
      subtitle = "CHLa values above 50 (n = 63) excluded for visualization purposes.",
      x = NULL,
      y = "CHLa (\u03BCg L\u207B\u00B9)"
    ) +
    theme_minimal() +
    theme(legend.position = "none",
          plot.subtitle = element_text(size = 10),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 10))    
  
  print(plot_chla)
  
    # ------------ CHLa log base 10 transformed -------------------------
        trnsfrm.data <- boxplot_data[["CHLa"]] %>%
          mutate(log10 = log10(yearly_mean))
  
          # Filter outliers above 2 to minimize y axis distortion
          trnsfrm.data <- trnsfrm.data %>%
            filter(log10 <= 2)
  
        plot_chlalog10 <- ggplot(trnsfrm.data, aes(x = interval, y = log10)) +
          geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
          geom_jitter(width = 0.1, alpha = 0.2, size = 0.7) +  
          geom_text(data = obs_counts[["CHLa"]], aes(x = interval, y = max(trnsfrm.data$log10, na.rm = TRUE), 
                                         label = paste("n =", n)), 
                  vjust = -0.5, size = 3, color = "black") +  # Add observation counts
          labs(
            title = "Chlorophyll-A (log\u2081\u2080 transformed)",
            subtitle = "log\u2081\u2080(CHLa) values above 2 (n = 11) excluded for visualization purposes.",
            x = NULL,
            y = "log\u2081\u2080(CHLa) (\u03BCg L\u207B\u00B9)"
          ) +
          theme_minimal() +
          theme(legend.position = "none",
                plot.subtitle = element_text(size = 10),
                axis.text.x = element_text(angle = 45, hjust = 1, size = 10))   
        
        print(plot_chlalog10)


# ------------- Total Nitrogen -------------------------
  ## Filter out TN values above 2 to minimize y axis distortion
  boxplot_data[["NITROGEN_TOTAL"]] <- boxplot_data[["NITROGEN_TOTAL"]] %>%
    filter(yearly_mean <= 2)
  
  plot_tn <- ggplot(boxplot_data[["NITROGEN_TOTAL"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.2, size = 0.7) +
    geom_text(data = obs_counts[["NITROGEN_TOTAL"]], 
              aes(x = interval, y = max(boxplot_data[["NITROGEN_TOTAL"]]$yearly_mean, na.rm = TRUE), 
                                                       label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  
    labs(
      title = "Total Nitrogen",
      subtitle = "TN values above 2 (n = 28) excluded for visualization purposes.",
      x = NULL,
      y = "TN (mg N L\u207B\u00B9)"
  ) +
    theme_minimal() +
    theme(legend.position = "none",
        plot.subtitle = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))   
  
  print(plot_tn)


# ------------- Total Phosphorus -------------------------
  ## Filter out TP values above .3 to minimize y axis distortion
  boxplot_data[["PHOSPHORUS_TOTAL"]] <- boxplot_data[["PHOSPHORUS_TOTAL"]] %>%
    filter(yearly_mean <= .3)
  
  plot_tp <- ggplot(boxplot_data[["PHOSPHORUS_TOTAL"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
    geom_jitter(width = 0.15, alpha = 0.2, size = 0.7) +
    geom_text(data = obs_counts[["PHOSPHORUS_TOTAL"]], 
              aes(x = interval, y = max(boxplot_data[["PHOSPHORUS_TOTAL"]]$yearly_mean, na.rm = TRUE), 
                                                       label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  
    labs(
      title = "Total Phosphorus",
      subtitle = "TP values above .3 (n = 21) excluded for visualization purposes.",
      x = NULL,
      y = "TP (mg N L\u207B\u00B9)"
  ) +
    theme_minimal() +
    theme(legend.position = "none",
        plot.subtitle = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))    
  
  print(plot_tp)

  
# ------------- Total Suspended Solids-------------------------
  ## Filter out TSS values above 50 (n = 34) to minimize y axis distortion
  boxplot_data[["TSS"]] <- boxplot_data[["TSS"]] %>%
    filter(yearly_mean <= 50)
  
  plot_tss <- ggplot(boxplot_data[["TSS"]], aes(x = interval, y = yearly_mean)) +
    geom_boxplot(aes(fill = interval), outlier.shape = NA) +  
    geom_jitter(width = 0.15, alpha = 0.2, size = 0.7) +
    geom_text(data = obs_counts[["TSS"]], 
              aes(x = interval, y = max(boxplot_data[["TSS"]]$yearly_mean, na.rm = TRUE), 
                                                       label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  
    labs(
      title = "Total Suspened Solids",
      subtitle = "TSS values above 50 (n = 6) excluded for visualization purposes.",
      x = NULL,
      y = "TSS (mg N L\u207B\u00B9)"
  ) +
    theme_minimal() +
    theme(legend.position = "none",
        plot.subtitle = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))    
  
  print(plot_tss)
 

```
### Continuous Axis for Boxplot Data

To overlay your MBLM trend lines onto the box plots in ggplot2, I need to align the categorical x-axis (the 5-year intervals) with the continuous x-axis (Year) used for the trend lines.

    Convert the Categorical Intervals to Numeric Midpoints: Assign numeric values (midpoints of the intervals) to each interval, effectively transforming your categorical x-axis into a continuous one.

    Update Data Frames: Add these midpoints to your boxplot_data so that each observation has an interval_midpoint.

    Adjust Plots: Modify your plotting code to use interval_midpoint as the x-axis, allowing both the box plots and trend lines to share the same continuous x-axis.

```{r}
# Create a data frame mapping intervals to midpoints
intervals_df <- data.frame(
  interval = c("1972-1974", "1975-1979", "1980-1984", "1985-1989",
               "1990-1994", "1995-1999", "2000-2004", "2005-2009",
               "2010-2014", "2015-2019", "2020-2023"),
  interval_midpoint = c(1973, 1977, 1982, 1987, 1992, 1997, 2002, 2007, 2012, 2017, 2021.5)
)

# Merge interval_midpoint into boxplot_data
for (var in names(boxplot_data)) {
  boxplot_data[[var]] <- boxplot_data[[var]] %>%
    left_join(intervals_df, by = "interval")
}

```


# MBLM Trend Estimates

"for the dataset that contains all the stations which have a min of 5 years of data, I would like to fit mblms to estimate the trend (slope) for CHLa, TN, TP, Sechi, TSS and DIN.  This is the same as what we did previously for individual stations, but now we are fitting a model for all the stations combined including those stations that met the 5 year min but not the 10 or 15 year min.  It would be helpful to see a plot of the data with the model."


## Derive May-Oct averages

Yearly means for each variable at each station where 5y or more of data (min 3 obs/yr). 

```{r, message=FALSE}
vars <- names(subset_data)

# List to store station averages for each variable
stationAvgs <- list()

# Loop by variable, as `subset_data` is structured by variable
for (var in vars) {
  
  # List to hold averages for the current variable
  station_avgs_current_var <- list()
  
  var_df <- subset_data[[var]]
  stations <- unique(var_df$STATION_ID)
  
  for (station in stations) {
    # Filter for current station and include only non-NA values of current var
    station_df <- var_df %>%
      filter(STATION_ID == station) %>%
      filter(!is.na(!!sym(var))) # dynamically reference the variable
             
    # Group observations by year and filter for only years with 3 or more obs
    # This is redundant, as the processed dataset generated in pt1 file already filters for min 3 obs per station-year. But no harm in being extra cautious.
      years_with_3_obs <- station_df %>%
        group_by(Year) %>%
        summarise(n = sum(!is.na(!!sym(var)))) %>%
        filter(n >= 3)
      
        # Calculate May-Oct average for the current station and variable, filtering to only include years with 3+ obs
        df_avg <- station_df %>%
          filter(Year %in% years_with_3_obs$Year) %>% # Only include years with >= 3 obs
          group_by(Year, STATION_ID) %>%
          summarise(MAY_OCT_AVG = mean(!!sym(var), na.rm = TRUE))

        # Add the May-Oct average for variable to the station_avgs list
        station_avgs_current_var[[station]] <- df_avg
  }
  
  # Only add station to the final list if it has valid averages for at least one variable
  if (length(station_avgs_current_var) > 0) {
    stationAvgs[[var]] <- station_avgs_current_var
  }
}

# Create data frame to save and share with Paul
avg_df_list <- list()

for (var in vars) {
  df <- bind_rows(stationAvgs[[var]])
  avg_df_list[[var]] <- df
}

all_vars_avg_df <- bind_rows(avg_df_list, .id = "variable")

# write_csv(all_vars_avg_df, "data/population_level_data/population_avg_data.csv")

```

## MBLM Model Fitting

```{r}
# Variable order within station average list
# [1] "SECCHI_DEPTH"     "CHLa"             "NITROGEN_TOTAL"  
# [4] "PHOSPHORUS_TOTAL" "TSS"              "DIN"

# Create function to loop through each variable and fit mblm
# Essential to use the stationAvgs list, as the boxplot_data lsit, despite having been derived from the same soruce and in the same sort of way, has had lots of outliers filtered out for plotting. Those outliers should still be in the mblm trend analysis.
mblm_func <- function(variable) {
  
  averages_list <- stationAvgs[[variable]]
  # Bind into single data frame for use in mblm
  df <- bind_rows(averages_list)
  
    # Run mblm model
    model <- mblm::mblm(MAY_OCT_AVG ~ Year, data = df)
    mod.sum <- mblm::summary.mblm(model)
    
    # Store results (directly, without creating the combined station-variable name)
    modelSummary <- list(
      slope = mod.sum$coefficients[2, 1],
      MAD = mod.sum$coefficients["Year", "MAD"],
      pvalue = mod.sum$coefficients["Year", 4],
      intercept = mod.sum$coefficients[1, 1]
    )
    
    modelSummaries <- modelSummary
  
  return(modelSummaries)
}


# ------- Run custom function in for loop -------

# Store population-level trend estimates for each variable in list
trends_results <- list()

  for (var in vars) {
  trends_results[[var]] <- mblm_func(var)
  }


```


## Extract Regression Statistics

```{r `regression stats`}
regStats<- list()

for (variable in names(trends_results)) {
      
  modelSums <- trends_results[[variable]]
  
      reg.df <- data.frame(variable = variable,
               model_slope = NA,
               model_MAD = NA,
               model_pval = NA,
               model_intercept = NA)
      
        p <- modelSums
      
        reg.df$model_slope <-  p$slope
        reg.df$model_MAD <-  p$MAD
        reg.df$model_pval <-  p$pvalue
        reg.df$model_intercept <-  p$intercept
      
      regStats[[variable]] <- reg.df
}
  

# ------- Create data frames from lists -------
regStats_allVars <- bind_rows(regStats) 
```


## Adjusted model intercept

Because the model intercept is based on x=0, and the x-axes are somewhere between 197X and 200x, I create an adjusted model intercept based on the following formula:

`adjusted_intercept = model_intercept + model_slope * min(Year)`

This requires first determining the minimum year for each variable at each station

```{r}
minYear <- list()

for (var in vars) {
  
  df <- bind_rows(stationAvgs[[var]])
  
  # Initialize the data frame for storing minimum years
  minYear.df <- data.frame(variable = var,
                           minYear = min(df$Year))
  
  # Store the data frame for this station
  minYear[[var]] <- minYear.df
}

minYear.df <- bind_rows(minYear)

```

Then using that data frame to calculate the adjusted intercept

```{r}
regressions_adjusted <- regStats_allVars %>%
  left_join(minYear.df, by = c("variable")) %>%
  mutate(adjusted_intercept = model_intercept + model_slope * minYear)

```

#### Plotting

```{r, echo=FALSE}
library(ggplot2)
# use regressions_adjusted along with the station averages list
# recall that, due to thow the output of the `mblm` function is structured, it cannot be passed to geom_smooth as a smoothing function. This results in having to 'manually' draw the trend lines.

# List to hold all plots 
all_plots_list <- list()
  
  for (current_var in names(stationAvgs)) {
    
    df_avgs <- stationAvgs[[current_var]] %>%
      bind_rows()
    
    df_reg <- regressions_adjusted %>%
      filter(variable == current_var)
    
    var_plot <- ggplot(df_avgs, aes(x = Year, y = MAY_OCT_AVG)) +
      geom_point(alpha = .3, position = position_jitter(width = 0.3)) +
        # Use annotate() instead of geom_segment() for the trend line. ggplot expects each aesthetic  to be mapped to a column in the data, and geom_segment() - which I originally used - was receiving values (x, xend, y, and yend) that had only a single length (min and max). Also, aes() is designed to map aesthetics to a SINGLE dataframe, whereas this plot requires reference to two dfs to produce the trend lines.Using `annotate` eliminates the warning because it explicitly adds a line based on single values (rather than having ggplot2 try to match them to rows in df_avgs)
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = 
                 df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               # calculate the y-value at the endpoint by adding the total change in y (slope * difference_in_years) to the intercept at the starting point
               color = "red", linewidth = .6, alpha = .8) +
      labs(title = current_var,
           subtitle = paste("Slope: ", round(df_reg$model_slope, 5),
                            "\np-value: ", round(df_reg$model_pval, 5)),
           x = "Year", y = NULL) +
      theme_minimal() +
      theme(plot.title = element_text(face = "bold"))

  all_plots_list[[current_var]] <- var_plot
}


# The plots are generated, now patch them together.
library(cowplot)

  # Use cowplot::plot_grid to combine plots 
  combined_plot <- plot_grid(plotlist = all_plots_list, ncol = 2)
  print(combined_plot)
  
  # ggsave("figures/population_trend_plots.jpg", combined_plot, height = 10, units = "in", dpi = 600)

```

# Overlay Trend Lines on Box Plots

```{r}
# Assuming you have 'df_reg' and 'df_avgs' for each variable
# Example for "SECCHI_DEPTH"

# Prepare data for trend line
df_avgs <- bind_rows(stationAvgs[["SECCHI_DEPTH"]]) %>% 
  filter(Year >= 1995) # Filter for years >= 1995

df_reg <- regressions_adjusted %>%
  filter(variable == "SECCHI_DEPTH")

# Generate a sequence of years for plotting the trend line
year_seq <- seq(min(df_avgs$Year), max(df_avgs$Year), by = 1)

# Compute predicted values from the model
trend_data <- data.frame(
  Year = year_seq,
  Trend = df_reg$model_intercept + df_reg$model_slope * year_seq
)

```

```{r}
# Plot for "SECCHI_DEPTH"
plot_secchi <- ggplot() +
  # Boxplots with continuous x-axis
  geom_boxplot(
    data = boxplot_data[["SECCHI_DEPTH"]] %>%
      filter(Year >= 1995),
    aes(x = interval_midpoint, y = yearly_mean, group = interval_midpoint),
    color = "black",
    width = 3.2,  # Adjust width as needed
    outlier.shape = NA
  ) +
  # Jittered points
  geom_jitter(
    data = boxplot_data[["SECCHI_DEPTH"]] %>%
      filter(Year >= 1995),
    aes(x = interval_midpoint, y = yearly_mean),
    width = 1, alpha = 0.2, size = 0.7
  ) +
  # Trend line
  geom_line(
    data = trend_data,
    aes(x = Year, y = Trend),
    color = "red", size = 1, alpha = 1
  ) +
  geom_text(data = obs_counts[["SECCHI_DEPTH"]], aes(x = interval_midpoint, y = max(boxplot_data[["SECCHI_DEPTH"]]$yearly_mean, na.rm = TRUE), 
                                   label = paste("n =", n)), 
            vjust = -0.5, size = 3, color = "black") +  # Add observation counts
  # Labels and themes
  labs(
    title = NULL,
    subtitle = paste("Slope:", round(df_reg$model_slope, 5), "p-value:", round(df_reg$model_pval, 5)),
    x = NULL,
    y = "Depth (m)"
  ) +
  # Adjust x-axis to display intervals
  scale_x_continuous(
    breaks = intervals_df$interval_midpoint,
    labels = intervals_df$interval
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9)
  )

print(plot_secchi)

 ggsave("figures/secchi_boxplot_trend_overlay.jpg", plot_secchi, height = 6, width = 8.2, units = "in", dpi = 600)

```



```{r}
## SAND BOX
## NOTE TO SELF: The reason there are so many fewer e.g. DIN station-years in the mblm model than in the  DIN vs TN plot is that we were not applying a year threshold to the DIN vs TN data set.

new <- all_vars_avg_df %>%
  mutate(station_yr_var = paste(STATION_ID, Year, variable, sep = "-"))
old <- read_csv("data/population_level_data/population_avg_data_ORIGINAL_error.csv") %>%
  mutate(station_yr_var = paste(STATION_ID, Year, variable, sep = "-"))

# Which station_yr_var vals not present in the old data set??
missing_in_old <- anti_join(new, old, by = "station_yr_var")


missing_in_old %>%
  mutate(MAY_OCT_AVG = as.character(MAY_OCT_AVG))

# PHOSPHORUS_TOTAL	1991	4AROA192.55	0.133333333333333	4AROA192.55-1991-PHOSPHORUS_TOTAL

unique(missing_in_old$STATION_ID)
```
