---
title: "Box plot/MBLM Trend Overlay"
output: html_document
date: "2024-10-09"
---

[This file is a copied and modified version of pt3_population_level_analysis
I wanted to keep that file intact, as the visualization request involves removing certain data points from the original dataset.]

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

"There is another thing I would like to try with the population data, but I'm not sure it will work.
I was wondering if you could take the model results from the plots you just sent, and overlay the trend lines on the box plots you made previously.  The aggregated data (box plots at 5-y intervals) are more useful for visualizing the data, but it would be helpful to also show how the model trend looks.
In doing this, I would like to re-run the mblm for Secchi and CHLa excluding the early data (1980-84), so the trend is based on data collected since 1995, and the model line added to the box plot starts with the 1995 group.
For TSS, I would like to fit the mblm to data spanning 1980 to 2005, which is where we have the most data (the model trend line would span the 5 boxes corresponding to this period).
For TN and TP, I would like to use all the data (i.e., the model you have already created)."

```{r, message = FALSE}
# Load the data
data <- read_csv("data/processed_dataset.csv")

  ## Remove 2024 obs since they are incomplete
  data <- data %>%
    filter(Year != 2024)
    
# This file contains a column with the number of years of observations. 
metadata <- read_csv("Data/station_metadata.csv")

```

# Subset Stations of Interest
 
```{r `subset data`}
# Function to subset data based on a minimum years of observations threshold. All station-variables in `data` have already been filtered to include only those with min 3 measurements per year (May-Oct). 
# In this case, 5 years of observations is the threshold.
subset_variable_data <- function(metadata_column, threshold = 5) {
  stations <- metadata %>%
    filter(!!sym(metadata_column) >= threshold) %>%
    pull(STATION_ID)
  
  return(data %>%
    filter(STATION_ID %in% stations))
}

# Create the subset_data list using the function
subset_data <- list(
  SECCHI_DEPTH = subset_variable_data("nYears_Secchi"),   
  CHLa = subset_variable_data("nYears_CHLA"),           
  NITROGEN_TOTAL = subset_variable_data("nYears_totNitrogen"), 
  PHOSPHORUS_TOTAL = subset_variable_data("nYears_totPhosphorus"), 
  TSS = subset_variable_data("nYears_TSS"),
  DIN = subset_variable_data("nYears_DIN")
)

```

### Outlier Removal & Subsetting

```{r}
## DIN outlier from station 4AROA192.94
din_data <- subset_data[["DIN"]]
index <- which.max(din_data$DIN[din_data$STATION_ID == "4AROA192.94"])
din_data$DIN[din_data$STATION_ID == "4AROA192.94"][index] <- NA
subset_data[["DIN"]] <- din_data

## CHLa outliers
### station 2-XDD000.40 (value = 1148) and station 5ASRN000.66 (value = 11592)
chla_data <- subset_data[["CHLa"]]
index1 <- which.max(chla_data$CHLa[chla_data$STATION_ID == "2-XDD000.40"])
index2 <- which.max(chla_data$CHLa[chla_data$STATION_ID == "5ASRN000.66"])
chla_data$CHLa[chla_data$STATION_ID == "2-XDD000.40"][index1] <- NA
chla_data$CHLa[chla_data$STATION_ID == "5ASRN000.66"][index2] <- NA
subset_data[["CHLa"]] <- chla_data


## Remove data from pre-1995 Secchi
subset_data[["SECCHI_DEPTH"]] <- subset_data[["SECCHI_DEPTH"]] %>%
  filter(Year >= 1995)

## Filter DIN to only include data up to 2004
subset_data[["DIN"]] <- subset_data[["DIN"]] %>%
  filter(Year <= 2004)

```


# Box-Whisker Plots

 Next, I would like to see box-whisker plots for each variable with the data partitioned into 5-y increments.  For convenience, I would suggest making the last box a 4-y increment (2020-23) and then go backwards in 5-y increments (2000-04, 2005-09, 2010-14, 2015-19, etc.).  As we go back past 2000, the data become more sparse.  *It would be helpful to indicate on the plot the number of observations used to make each box.  We will likely adopt a cut-off below which we will not include that interval/variable, but I don't know offhand what that should be.*


### Partition Data - 5y Increments

Working backwards from 2023, with first increment (2019-2023) being 4 years.

```{r}
create_intervals <- function(year) {
  case_when(
    year >= 2020 ~ "2020-23",
    year >= 2015 ~ "2015-19",
    year >= 2010 ~ "2010-14",
    year >= 2005 ~ "2005-09",
    year >= 2000 ~ "2000-04",
    year >= 1995 ~ "1995-99",
    year >= 1990 ~ "1990-94",
    year >= 1985 ~ "1985-89",
    year >= 1980 ~ "1980-84",
    year >= 1975 ~ "1975-79",
    year >= 1972 ~ "1972-74",
    TRUE ~ NA
  )
}



# Derive May-Oct avgs for each station-year
# Use function to assign each May-Oct avg to an interval, by variable
boxplot_data <- list()
for (var in names(subset_data)) {
  x <- subset_data[[var]] %>%
    filter(!is.na(!!sym(var))) 
  min_year <- min(x$Year)
  
  x_w_intervals <- x %>%
    mutate(interval = create_intervals(Year))
  
  boxplot_data[[var]] <- x_w_intervals %>%
        group_by(STATION_ID, Year) %>%
        summarize(yearly_mean = mean(!!sym(var), na.rm = TRUE),
                  interval = first(interval))

}

# Create log10 transformed CHLA data for plotting
boxplot_data[["CHLa_log10"]] <- boxplot_data[["CHLa"]] %>%
  mutate(yearly_mean = log10(yearly_mean ))

```

### Continuous Axis for Boxplot Data

To overlay your MBLM trend lines onto the box plots in ggplot2, I need to align the categorical x-axis (the 5-year intervals) with the continuous x-axis (Year) used for the trend lines.

    Convert the Categorical Intervals to Numeric Midpoints: Assign numeric values (midpoints of the intervals) to each interval, effectively transforming your categorical x-axis into a continuous one.

    Update Data Frames: Add these midpoints to your boxplot_data so that each observation has an interval_midpoint.

    Adjust Plots: Modify your plotting code to use interval_midpoint as the x-axis, allowing both the box plots and trend lines to share the same continuous x-axis.

```{r}
# Create a data frame mapping intervals to midpoints
intervals_df <- data.frame(
  interval = c("1972-74", "1975-79", "1980-84", "1985-89",
               "1990-94", "1995-99", "2000-04", "2005-09",
               "2010-14", "2015-19", "2020-23"),
  interval_midpoint = c(1973, 1977, 1982, 1987, 1992, 1997, 2002, 2007, 2012, 2017, 2021.5)
)

# Merge interval_midpoint into boxplot_data
for (var in names(boxplot_data)) {
  boxplot_data[[var]] <- boxplot_data[[var]] %>%
    left_join(intervals_df, by = "interval")
}

```


How many observations per interval, per variable? This information needs to be incorporated into the plots.

```{r}
# List of Nobs per interval for incorporation into plots
obs_counts <- list()
for (var in names(boxplot_data)) {
  x <- boxplot_data[[var]]
  x_counts <- x %>%
    group_by(interval, interval_midpoint) %>%
    summarise(n = n())
  
  obs_counts[[var]] <- x_counts
}

```



# MBLM Trend Estimates

"for the dataset that contains all the stations which have a min of 5 years of data, I would like to fit mblms to estimate the trend (slope) for CHLa, TN, TP, Sechi, TSS and DIN.  This is the same as what we did previously for individual stations, but now we are fitting a model for all the stations combined including those stations that met the 5 year min but not the 10 or 15 year min.  It would be helpful to see a plot of the data with the model."


## Derive May-Oct averages

Yearly means for each variable at each station where 5y or more of data (min 3 obs/yr). 

```{r, message=FALSE}
vars <- names(subset_data)


## Second round of subsetting, now that boxplot data has been created. This allows the box plots for these varaibles to include all intervals, while the trend line will be based on a restricted subset, and thus only span the middle intervals where data is most robust.
## Filter to only include 1980-2005 for TSS
subset_data[["TSS"]] <- subset_data[["TSS"]] %>%
  filter(Year >= 1980, Year <= 2004)

## Remove data pre-1990 for CHLa
subset_data[["CHLa"]] <- subset_data[["CHLa"]] %>%
  filter(Year >= 1995)



# List to store station averages for each variable
stationAvgs <- list()

# Loop by variable, as `subset_data` is structured by variable
for (var in vars) {
  
  # List to hold averages for the current variable
  station_avgs_current_var <- list()
  
  var_df <- subset_data[[var]]
  stations <- unique(var_df$STATION_ID)
  
  for (station in stations) {
    # Filter for current station and include only non-NA values of current var
    station_df <- var_df %>%
      filter(STATION_ID == station) %>%
      filter(!is.na(!!sym(var))) # dynamically reference the variable
             
    # Group observations by year and filter for only years with 3 or more obs
    # This is redundant, as the processed dataset generated in pt1 file already filters for min 3 obs per station-year. But no harm in being extra cautious.
      years_with_3_obs <- station_df %>%
        group_by(Year) %>%
        summarise(n = sum(!is.na(!!sym(var)))) %>%
        filter(n >= 3)
      
        # Calculate May-Oct average for the current station and variable, filtering to only include years with 3+ obs
        df_avg <- station_df %>%
          filter(Year %in% years_with_3_obs$Year) %>% # Only include years with >= 3 obs
          group_by(Year, STATION_ID) %>%
          summarise(MAY_OCT_AVG = mean(!!sym(var), na.rm = TRUE))

        # Add the May-Oct average for variable to the station_avgs list
        station_avgs_current_var[[station]] <- df_avg
  }
  
  # Only add station to the final list if it has valid averages for at least one variable
  if (length(station_avgs_current_var) > 0) {
    stationAvgs[[var]] <- station_avgs_current_var
  }
}



avg_df_list <- list()

for (var in vars) {
  df <- bind_rows(stationAvgs[[var]])
  avg_df_list[[var]] <- df
}

## CHLA log10 transformed
avg_df_list[["CHLa_log10"]] <- avg_df_list[["CHLa"]] %>%
  mutate(MAY_OCT_AVG = log10(MAY_OCT_AVG))



```

## MBLM Model Fitting

```{r}
# Variable order within station average list
# [1] "SECCHI_DEPTH"     "CHLa"             "NITROGEN_TOTAL"  
# [4] "PHOSPHORUS_TOTAL" "TSS"              "DIN"

# Create function to loop through each variable and fit mblm
# Essential to use the stationAvgs list, as the boxplot_data lsit, despite having been derived from the same soruce and in the same sort of way, has had lots of outliers filtered out for plotting. Those outliers should still be in the mblm trend analysis.
mblm_func <- function(variable) {
  
  df <- avg_df_list[[variable]]
  
    # Run mblm model
    model <- mblm::mblm(MAY_OCT_AVG ~ Year, data = df)
    mod.sum <- mblm::summary.mblm(model)
    
    # Store results (directly, without creating the combined station-variable name)
    modelSummary <- list(
      slope = mod.sum$coefficients[2, 1],
      MAD = mod.sum$coefficients["Year", "MAD"],
      pvalue = mod.sum$coefficients["Year", 4],
      intercept = mod.sum$coefficients[1, 1]
    )
    
    modelSummaries <- modelSummary
  
  return(modelSummaries)
}


# ------- Run custom function in for loop -------
vars <- names(avg_df_list)

# Store population-level trend estimates for each variable in list
trends_results <- list()

  for (var in vars) {
  trends_results[[var]] <- mblm_func(var)
  }

```


## Extract Regression Statistics

```{r `regression stats`}
regStats<- list()

for (variable in names(trends_results)) {
      
  modelSums <- trends_results[[variable]]
  
      reg.df <- data.frame(variable = variable,
               model_slope = NA,
               model_MAD = NA,
               model_pval = NA,
               model_intercept = NA)
      
        p <- modelSums
      
        reg.df$model_slope <-  p$slope
        reg.df$model_MAD <-  p$MAD
        reg.df$model_pval <-  p$pvalue
        reg.df$model_intercept <-  p$intercept
      
      regStats[[variable]] <- reg.df
}
  

# ------- Create data frames from lists -------
regStats_allVars <- bind_rows(regStats) 
```


## Adjusted model intercept

Because the model intercept is based on x=0, and the x-axes are somewhere between 197X and 200x, I create an adjusted model intercept based on the following formula:

`adjusted_intercept = model_intercept + model_slope * min(Year)`

This requires first determining the minimum year for each variable at each station...

```{r}
minYear <- list()

for (var in vars) {
  
  df <- avg_df_list[[var]]
  
  # Initialize the data frame for storing minimum years
  minYear.df <- data.frame(variable = var,
                           minYear = min(df$Year))
  
  # Store the data frame for this station
  minYear[[var]] <- minYear.df
}

minYear.df <- bind_rows(minYear)

```

...then using that data frame to calculate the adjusted intercept

```{r}
regressions_adjusted <- regStats_allVars %>%
  left_join(minYear.df, by = c("variable")) %>%
  mutate(adjusted_intercept = model_intercept + model_slope * minYear)


### NOTE 11/8/2024  -- Data included in the models was modified per Paul's email on 11/7. Thus, reg. stats for several of the variables do not match those I sent to him on 11/5 (e.g., DIN now only includes data through 2004; TSS data is now 1980-2005, etc.)

```


# Overlay Trend Lines on Box Plots

```{r}
library(ggtext)
# Create named list for defining y axis labels in plots
y_titles <- list(
  SECCHI_DEPTH = "Depth (m)",
  CHLa = "CHLa (\u03BCg L\u207B\u00B9)",
  NITROGEN_TOTAL = "TN (mg N L\u207B\u00B9)",
  PHOSPHORUS_TOTAL = "TP (mg P L\u207B\u00B9)",
  TSS = "TSS (mg L\u207B\u00B9)",
  DIN = "DIN (mg N L\u207B\u00B9)",
  CHLa_log10 = "log\u2081\u2080(CHLa) (\u03BCg L\u207B\u00B9)"
)


## !----------------- Custom plotting function -----------------! ##
boxtrend_overlay <- function(current_var) {
  
  ## y limits for visualization
  y_lims <- list(
    SECCHI_DEPTH = 8.2,
    CHLa = 125,
    NITROGEN_TOTAL = 2,
    PHOSPHORUS_TOTAL = .25,
    TSS = 40,
    DIN = 2,
    CHLa_log10 = 2
  )
  
  # Calculate trend line data within the plotting function
  df_avgs <- avg_df_list[[current_var]]
  df_reg <- regressions_adjusted %>%
    filter(variable == current_var)
  
  # Default start and end y-values for the trend line
  start_y <- df_reg$adjusted_intercept
  end_y <- df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year))
  
  # Calculate default x min and x max based on box plot edges
  min_x <- min(boxplot_data[[current_var]]$interval_midpoint) - (3.2 / 2)
  max_x <- max(boxplot_data[[current_var]]$interval_midpoint) + (3.2 / 2)
  
  # Special cases for CHLa_log10 and TSS
  if (current_var == "CHLa_log10") {
    start_x <- 1997 - (3.2 / 2) 
    end_x <- max_x  # End at the default max
    start_y <- df_reg$model_slope * (start_x - min(df_avgs$Year)) + df_reg$adjusted_intercept
  } else if (current_var == "TSS") {
    start_x <- 1982 - (3.2 / 2)
    end_x <- 2002 + (3.2 / 2)
    start_y <- df_reg$model_slope * (start_x - min(df_avgs$Year)) + df_reg$adjusted_intercept
    end_y <- df_reg$model_slope * (end_x - min(df_avgs$Year)) + df_reg$adjusted_intercept
  } else {
    # Use default values for all other variables
    start_x <- min_x
    end_x <- max_x
  }
  
  obs_counts[[current_var]]$label <- paste0("*", obs_counts[[current_var]]$n, "*")
  
  # Generate plot
  plot <- ggplot() +
    # Boxplots with continuous x-axis
    geom_boxplot(
      data = boxplot_data[[current_var]],
      aes(x = interval_midpoint, y = yearly_mean, group = interval_midpoint),
      color = "black",
      width = 3.2,  # Adjust width as needed
      outlier.shape = NA
    ) +
    # Jittered points
    geom_jitter(
      data = boxplot_data[[current_var]],
      aes(x = interval_midpoint, y = yearly_mean),
      width = 1, alpha = 0.3, size = 0.7
    ) +
    # Annotated trend line spanning from start_x to end_x with respective y values
    annotate(
      "segment",
      x = start_x,
      xend = end_x,
      y = start_y,
      yend = end_y,
      color = "red", linewidth = .85, alpha = 1
    ) +
    geom_text(
      data = obs_counts[[current_var]],
      aes(x = interval_midpoint, y = y_lims[[current_var]], label = n),
      vjust = -0.5, size = 3.7, color = "black"
    ) +
    labs(
      title = NULL,
      x = NULL,
      y = y_titles[[current_var]]
    ) +
    # Adjust x-axis to display intervals
    scale_x_continuous(
      breaks = intervals_df$interval_midpoint,
      labels = intervals_df$interval
    ) +
    scale_y_continuous(limits = c(0, y_lims[[current_var]])) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 0, hjust = 0.5, size = 11),
      axis.text.y = element_text(size = 11),
      axis.line = element_line(color = "black"),
      axis.title.y = element_text(size = 12)
    )
  
  return(plot)
}

```


## Generate Plots

```{r}
# Create list of plots for each variable
plots_list <- list()

for (var in vars) {
  plots_list[[var]] <- boxtrend_overlay(var)
  
  ggsave(paste0("figures/boxplot_trend_overlay_", var, ".jpg"), plots_list[[var]], width = 10, height = 7, units = "in", dpi = 1000)
}

plots_list

```

```{r}
## OUTLIER EVALUATION

# For TN, you had previously used a range up to 2 mg/L, which looked good.  But could you also check on the 4 data points that are > 4 mg/L.  If any of these are cases where there is a single monthly value that is unrealistically high (>5 mg/L) then that value should be dropped and the May-Oct average value re-calculated. 

avg_df_list[["NITROGEN_TOTAL"]] %>%
  filter(MAY_OCT_AVG > 3)
## all 4 vals above 4 are from the same station, 1BXEF000.23	 

yrs_interest <- c(2009, 2010, 2015, 2016)
x <- subset_data[["NITROGEN_TOTAL"]] %>%
  filter(STATION_ID == "1BXEF000.23", Year %in% yrs_interest) %>%
  select(Year, Month, NITROGEN_TOTAL)

hist(x$NITROGEN_TOTAL,
     main = "TN: Station 1BXEF000.23",
     xlab = "TN mg/L")

## For TP, you had previously used a range up to 0.3 mg/L, which looked good.  Here too, could you check on the 3 data points that are >1 mg/L to make sure these are not the result of a single monthly outlier.
TP_above1  <- avg_df_list[["PHOSPHORUS_TOTAL"]] %>%
  filter(MAY_OCT_AVG > 1)

a <- subset_data[["PHOSPHORUS_TOTAL"]] %>%
  filter(STATION_ID == "2-XDD000.40", Year == 2007) %>%
  select(STATION_ID, Year, Month, PHOSPHORUS_TOTAL)
b <- subset_data[["PHOSPHORUS_TOTAL"]] %>%
  filter(STATION_ID == "5ASRN000.66", Year %in% c(2004, 2008, 2010)) %>%
  select(STATION_ID, Year, Month, PHOSPHORUS_TOTAL)

bind_rows(a, b)

```
