---
title: "Lake Average Trends"
output: html_document
---

This file derives model trends for four lakes - Claytor, Anna, Kerr and SML - using average values across all stations within each lake. 

Steps:

    derive the May-Oct average values for each station by year, 
    average these across stations within each year
    run a mblm on the lake average values
    generate output csv with results

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Prepare Data

```{r, message = FALSE}
# Load the data
data <- read_csv("data/processed_dataset.csv")
stations.df <- read_csv("data/lake_averages/lakeAvgStations_list.csv")
stations <- unique(stations.df$STATION_ID)
lakes <- unique(stations.df$Lake)

# Subset data to include only relevant stations
data.sub <- data %>% 
  filter(STATION_ID %in% stations.df$STATION_ID) %>%
  left_join(stations.df, by = "STATION_ID")

# Remove DIN outlier from SML in 1983
index <- which.max(data.sub$DIN[data.sub$Lake == "SML" & data.sub$Year == 1983])
data.sub$DIN[data.sub$Lake == "SML" & data.sub$Year == 1983][index] <- NA

# Define vector for looping through variables
vars <- c("CHLa", "PHOSPHORUS_TOTAL", "NITROGEN_TOTAL", "DIN", "TSS", "SECCHI_DEPTH")

```

# Derive May-Oct Averages

```{r, message = FALSE}
stationAvgs <- list()

## Resulting list structure:
## list[[station]]$variable  variable = df with Year, STATION_ID, MAY_OCT_AVG
for (station in stations) {
  
  station_avgs <- list()
  
  for (variable in vars) {
    # Filter for current station and include only non-NA values of current var
    data_station_variable <- data.sub %>%
      filter(STATION_ID == station) %>%
      filter(!is.na(!!sym(variable))) # dynamically reference the variable

    # Confirm at least 10 yrs of data
    if (length(unique(data_station_variable$Year)) >= 10) {
      # Determine which years have at least 3 obs/yr
      years_with_3_obs <- data_station_variable %>%
        group_by(Year) %>%
        summarise(n = sum(!is.na(!!sym(variable)))) %>%
        filter(n >= 3)

      if (nrow(years_with_3_obs) >= 10) {
        # Calculate May-Oct average for the current station and variable
        the_averages <- data_station_variable %>%
          filter(Year %in% years_with_3_obs$Year) %>% # Only include years with >= 3 obs
          group_by(Year, STATION_ID) %>%
          summarise(MAY_OCT_AVG = mean(!!sym(variable), na.rm = TRUE),
                    Lake = first(Lake))

        # Add the May-Oct average for variable to the station_avgs list
        station_avgs[[variable]] <- the_averages
      }
    }
  }
  
  # Only add station to the final list if it has valid averages for at least one variable
  if (length(station_avgs) > 0) {
    stationAvgs[[station]] <- station_avgs
  }
}

# Simplify the list structure
bound_averages <- list()
for (station in stations) {
  bound <- bind_rows(stationAvgs[[station]], .id = "variable")
  bound_averages[[station]] <- bound
}

# Combine all station averages into one dataframe
mayOctAvgs_df <- bind_rows(bound_averages)

```

# Derive Averages by Lake

```{r}
lakeAvgs <- mayOctAvgs_df %>%
  group_by(Lake, Year, variable) %>%
  summarize(LakeAvg = mean(MAY_OCT_AVG, na.rm = TRUE)) %>%
  ungroup()

write_csv(lakeAvgs, "data/lake_averages/lakeAverages_byVariable_Year.csv")

```

# MBLM Trends

### Custom function to derive trends

```{r, warning = FALSE}
trends_func <- function(lake, var) {
  
  modelSummary <- list()  
  
  df <- lakeAvgs %>%
    filter(Lake == lake, variable == var)
  
  # Remove NA values -- there should not be any, but just in case.
  df <- df %>%
    filter(!is.na(LakeAvg)) %>%
    ungroup()
  
  # Run mblm model
  model <- mblm::mblm(LakeAvg ~ Year, data = df)
  mod.sum <- mblm::summary.mblm(model)
          
  # Store results (directly, without creating the combined station-variable name)
  modelSummary <- list(
    slope = mod.sum$coefficients[2, 1],
    MAD = mod.sum$coefficients["Year", "MAD"],
    pvalue = mod.sum$coefficients["Year", 4],
    intercept = mod.sum$coefficients[1, 1]
  )
  
  return(modelSummary)  # Return the results
}

# ------- Run custom function in for loop -------
## list[[station]]$variable  variable = list with slope, MAD, pval, intercept

# Trends for each lake by variable
trends_results <- list()

for (l in lakes) {
  lake_vars <- unique(lakeAvgs[lakeAvgs$Lake == l,]$variable)
  p <- vars[vars %in% lake_vars]
  
  for (var in p) {
      trends_results[[l]][[var]] <- trends_func(l, var)
  }
}

```


### Extract Regression Statistics into single table

```{r `regression stats`}
regression_statistics <- list()

for (lake in names(trends_results)) {
      
  modelSums <- trends_results[[lake]]
  
      reg.df <- data.frame(Lake = lake,
                           variable = vars[vars %in% names(modelSums)],
               model_slope = rep(NA, length(trends_results[[lake]])),
               model_MAD = rep(NA, length(trends_results[[lake]])),
               model_pval = rep(NA, length(trends_results[[lake]])),
               model_intercept = rep(NA, length(trends_results[[lake]]))
      )
      
  for (i in seq_along(trends_results[[lake]])) {
      
        p <- modelSums[[i]]
      
        reg.df$model_slope[i] <-  p$slope
        reg.df$model_MAD[i] <-  p$MAD
        reg.df$model_pval[i] <-  p$pvalue
        reg.df$model_intercept[i] <-  p$intercept
      }
      
      regression_statistics[[lake]] <- reg.df
}
  

# ------- Create data frames from lists -------
lakeAvgs_regressionStats <- bind_rows(regression_statistics) 

write_csv(lakeAvgs_regressionStats, "data/lake_averages/lakeAverage_regressionStats.csv")

```

# Visualize Trends

I will create figures of gridded plots as I did for select stations in the pt. 2 file. Rather than by station, each figure will be by lake. 

```{r, message = FALSE}
# `cowplot` package for use in arranging the plots into single figures
library(cowplot)
```

### Adjusted model intercept

Because the model intercept is based on x=0, and the x-axes are somewhere between 1980 and 200x, I created an adjusted model intercept based on the following formula:

`adjusted_intercept = model_intercept + model_slope * min(Year)`

This required first determining the minimum year for each variable at each station

```{r}
minYear <- list()

for (lake in unique(lakeAvgs_regressionStats$Lake)) {
  # Filter variables that exist in the current station
  station_vars <- vars[vars %in% (lakeAvgs_regressionStats %>%
                         filter(Lake == lake) %>%
                         pull(variable))]

  # Initialize the data frame for storing minimum years
  minYear.df <- data.frame(Lake = lake,
                           variable = station_vars,
                           minYear = rep(NA, length(station_vars)),
                           maxYear = rep(NA, length(station_vars))
                           )
  
  # Loop through the variables and calculate the minimum year
  for (i in seq_along(station_vars)) {
    var <- station_vars[i]
    minYear.df$minYear[i] <- min(lakeAvgs %>%
                                  filter(Lake == lake, variable == var) %>%
                                  pull(Year))
    minYear.df$maxYear[i] <- max(lakeAvgs %>%
                                  filter(Lake == lake, variable == var) %>%
                                  pull(Year))
  }
  
  # Store the data frame for this station
  minYear[[lake]] <- minYear.df
}

minYear_df <- bind_rows(minYear)

```

Now use  `minYear.df` to calculate the adjusted intercept

```{r}
regressions_adjusted <- lakeAvgs_regressionStats %>%
  left_join(minYear_df, by = c("Lake", "variable")) %>%
  mutate(adjusted_intercept = model_intercept + model_slope * minYear)

```



Additional fields can be added to the data frame to facilitate the plotting process. This includes a field for defining p-val rendering and a field for y axis labeling.

The regression statistics still need adjusted y-intercepts, as well.

```{r}
regressions_adjusted <- regressions_adjusted %>%
  mutate(pval_render = case_when(
    model_pval > 0.05 ~ paste0("p = ", formatC(model_pval, digits = 2, format = "f")),
    model_pval < 0.05 & model_pval >= 0.001 ~ paste0("p = ", formatC(model_pval, digits = 3, format = "f")),
    model_pval < 0.001 ~ "p < 0.001"
  ),
  y_label = case_when(
    variable == "CHLa" ~ "CHLa (\u03BCg L\u207B\u00B9)",
    variable == "PHOSPHORUS_TOTAL" ~ "TP (mg L\u207B\u00B9)",
    variable == "NITROGEN_TOTAL" ~ "TN (mg L\u207B\u00B9)",
    variable == "TSS" ~ "TSS (mg L\u207B\u00B9)",
    variable == "SECCHI_DEPTH" ~ "Secchi Depth (m)",
    variable == "DIN" ~ "DIN (mg L\u207B\u00B9)"
  ))

```


### a. Claytor Lake

```{r, fig.width=7, fig.height=7}
lake <- "Claytor Lake"
data_claytor <- regressions_adjusted %>%
  filter(Lake == lake)

# List to hold the plots for this station
claytor_list <- list()

# Set a standard x-axis range across all plots (adjust as per data)
min_year <- 1973
max_year <- 2025  # Set the max year based on your data


# ! ---- Plot Generation ----- ! #
for (current_var in data_claytor$variable)  {
  df_avgs <- lakeAvgs %>%
    filter(Lake == lake,
           variable == current_var)
  df_reg <- data_claytor %>%
    filter(variable == current_var)
  
  # Extract the y_label for the current variable
  current_y_label <- df_reg$y_label[1]
  # max y value for dynamic range -- needed, since the ylim needs to be set at 0 on the low end
  max_y <- max(df_avgs$LakeAvg, na.rm = TRUE) * 1.2
  
  # Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(df_avgs$Year) - 5  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis
  
  var_plot <- ggplot(df_avgs, aes(x = Year, y = LakeAvg)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = df_reg$pval_render, size = 3.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x = NULL, y = current_y_label) +  # Use the extracted y_label
      scale_y_continuous(limits = c(0, max_y), labels = function(y) ifelse(y == 0, "0", format(y, digits = 1, nsmall = 0))) +
      # Set x-axis limits to be consistent
      scale_x_continuous(limits = c(min_year, max_year)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 9.5),        # Adjust axis title size if needed
        axis.text = element_text(size = 7.5, color = "black"),         # Adjust axis text size
        plot.subtitle = element_text(size = 9),     # Adjust subtitle size
        plot.margin = unit(c(.3, .3, .3, .3), "cm")
      )
  
  claytor_list[[current_var]] <- var_plot
}

  
  claytorAvgs_plot <- plot_grid(plotlist = claytor_list, ncol = 2,
                              vjust = 4)
  print(claytorAvgs_plot)
  
ggsave("claytorAvgs_plot.jpg", width = 7.2, height = 7, units = "in", dpi = 1000)

```

### b. Smith Mountain Lake

```{r, fig.width=7, fig.height=7}
lake <- "SML"
data_SML <- regressions_adjusted %>%
  filter(Lake == lake)

# List to hold the plots for this station
SML_list <- list()

# Set a standard x-axis range across all plots (adjust as per data)
min_year <- minYear_df %>%
  filter(Lake == lake) %>%
  pull(minYear) %>%
  min()
max_year <- 2025  # Set the max year based on your data


# ! ---- Plot Generation ----- ! #
for (current_var in data_SML$variable)  {
  df_avgs <- lakeAvgs %>%
    filter(Lake == lake,
           variable == current_var)
  df_reg <- data_SML %>%
    filter(variable == current_var)
  
  # Extract the y_label for the current variable
  current_y_label <- df_reg$y_label[1]
  # max y value for dynamic range -- needed, since the ylim needs to be set at 0 on the low end
  max_y <- max(df_avgs$LakeAvg, na.rm = TRUE) * 1.2
  
  # Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(df_avgs$Year) - 5  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis
  
  var_plot <- ggplot(df_avgs, aes(x = Year, y = LakeAvg)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = df_reg$pval_render, size = 3.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x = NULL, y = current_y_label) +  # Use the extracted y_label
      scale_y_continuous(limits = c(0, max_y), labels = function(y) ifelse(y == 0, "0", format(y, digits = 1, nsmall = 0))) +
      # Set x-axis limits to be consistent
      scale_x_continuous(limits = c(min_year, max_year)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 9.5),        # Adjust axis title size if needed
        axis.text = element_text(size = 7.5, color = "black"),         # Adjust axis text size
        plot.subtitle = element_text(size = 9),     # Adjust subtitle size
        plot.margin = unit(c(.3, .3, .3, .3), "cm")
      )
  
  SML_list[[current_var]] <- var_plot
}

  
  SMLAvgs_plot <- plot_grid(plotlist = SML_list, ncol = 2,
                              vjust = 4)
  print(SMLAvgs_plot)
  
ggsave("SMLAvgs_plot.jpg", width = 7.2, height = 7, units = "in", dpi = 1000)

```

### c. Kerr Reservoir

```{r, fig.width=8, fig.height=7}
lake <- "Kerr Reservoir"
data_kerr <- regressions_adjusted %>%
  filter(Lake == lake)

# List to hold the plots for this station
kerr_list <- list()

# Set a standard x-axis range across all plots (adjust as per data)
min_year <- minYear_df %>%
  filter(Lake == lake) %>%
  pull(minYear) %>%
  min()
max_year <- 2025  # Set the max year based on your data


# ! ---- Plot Generation ----- ! #
for (current_var in data_kerr$variable)  {
  df_avgs <- lakeAvgs %>%
    filter(Lake == lake,
           variable == current_var)
  df_reg <- data_kerr %>%
    filter(variable == current_var)
  
  # Extract the y_label for the current variable
  current_y_label <- df_reg$y_label[1]
  # max y value for dynamic range -- needed, since the ylim needs to be set at 0 on the low end
  max_y <- max(df_avgs$LakeAvg, na.rm = TRUE) * 1.2
  
  # Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(df_avgs$Year) - 5  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis

  
  var_plot <- ggplot(df_avgs, aes(x = Year, y = LakeAvg)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = df_reg$pval_render, size = 3.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x = NULL, y = current_y_label) +  # Use the extracted y_label
      scale_y_continuous(limits = c(0, max_y), 
                         labels = function(y) ifelse(y == 0, "0", format(y, digits = 1, nsmall = 0))
                         ) +
      # Set x-axis limits to be consistent
      scale_x_continuous(limits = c(min_year, max_year)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 9.5),        # Adjust axis title size if needed
        axis.text = element_text(size = 7.5, color = "black"),         # Adjust axis text size
        plot.subtitle = element_text(size = 9),
        plot.margin = unit(c(.3, .3, .3, .3), "cm")
      )
  
  kerr_list[[current_var]] <- var_plot
}

  
  kerrAvgs_plot <- plot_grid(plotlist = kerr_list, ncol = 2,
                              vjust = 4)
  print(kerrAvgs_plot)
  
ggsave("kerrAvgs_plot.jpg", width = 7.2, height = 7, units = "in", dpi = 1000)

```

### d. Lake Anna

```{r, fig.width=7, fig.height=7}
lake <- "Lake Anna"
data_anna <- regressions_adjusted %>%
  filter(Lake == lake)

# List to hold the plots for this station
anna_list <- list()

# Set a standard x-axis range across all plots (adjust as per data)
min_year <- minYear_df %>%
  filter(Lake == lake) %>%
  pull(minYear) %>%
  min()
max_year <- 2025  # Set the max year based on your data


# ! ---- Plot Generation ----- ! #
for (current_var in data_anna$variable)  {
  df_avgs <- lakeAvgs %>%
    filter(Lake == lake,
           variable == current_var)
  df_reg <- data_anna %>%
    filter(variable == current_var)
  
  # Extract the y_label for the current variable
  current_y_label <- df_reg$y_label[1]
  # max y value for dynamic range -- needed, since the ylim needs to be set at 0 on the low end
  max_y <- max(df_avgs$LakeAvg, na.rm = TRUE) * 1.2
  
  # Calculate position for p-value: place it dynamically near the upper right, clear of the data points
  pval_x_pos <- max(df_avgs$Year) - 5  # Adjust the 5 to change how far left from the max Year
  pval_y_pos <- max_y * 0.95           # Place it slightly below the upper limit of the y-axis
  
  var_plot <- ggplot(df_avgs, aes(x = Year, y = LakeAvg)) +
    geom_point() +
      # Add p-value inside the plot, positioning dynamically
      annotate("text", x = pval_x_pos, y = pval_y_pos, 
               label = df_reg$pval_render, size = 3.2, hjust = 1) +
      # Regression line
      annotate("segment", x = min(df_avgs$Year), 
               xend = max(df_avgs$Year), 
               y = df_reg$adjusted_intercept, 
               yend = df_reg$adjusted_intercept + df_reg$model_slope * (max(df_avgs$Year) - min(df_avgs$Year)),
               color = "red", linewidth = .42, alpha = .6) +
      labs(title = NULL,
           x = NULL, y = current_y_label) +  # Use the extracted y_label
      scale_y_continuous(limits = c(0, max_y), labels = function(y) ifelse(y == 0, "0", format(y, digits = 1, nsmall = 0))) +
      # Set x-axis limits to be consistent
      scale_x_continuous(limits = c(min_year, max_year)) +
      theme_minimal() +
      # Make x and y axes black without adding a border around the entire plot
      theme(
        axis.line = element_line(color = "black"),  # Make x and y axes black
        axis.title = element_text(size = 9.5),        # Adjust axis title size if needed
        axis.text = element_text(size = 7.5, color = "black"),         # Adjust axis text size
        plot.subtitle = element_text(size = 9),     # Adjust subtitle size
        plot.margin = unit(c(.3, .3, .3, .3), "cm")
      )
  
  anna_list[[current_var]] <- var_plot
}

  
  annaAvgs_plot <- plot_grid(plotlist = anna_list, ncol = 2,
                              vjust = 4)
  print(annaAvgs_plot)
  
ggsave("LakeAnnaAvgs_plot.jpg", width = 7.2, height = 7, units = "in", dpi = 1000)

```

