---
title: "Publication Plot: Station-Level Long Term Trends"
author: "Andrew Cameron"
date: "2024-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

```{r, message = FALSE}
# Load the data
data <- read_csv("data/processed_dataset.csv")
trends <- read_csv("data/station_level_data/longTerm_Annual_Regression_Statistics_ALLSTATIONS.csv")
# This file contains a column with the number of years of observations. 
metadata <- read_csv("Data/station_metadata.csv")

```

# Plot Creation

Objective: Panelled plot with 3 variables per row; 6 vars total (Secchi, CHLa, TSS, Total Nitrogen, Total Phosphorus, DIN). 
Each panel will have trend lines for all relevant stations. 
Lines should be colored based on the significance of the slope (significant positive, significant negative, not significant). 

Plotting mblm trend lines of this sort requires a data frame with the min and max years of observations for each station-variable combination to be able to define the x start and end points. Adjusted model slopes are also required, as model slopes are based on x=0 but our x ranges from 1970s to 20xx. 

`adjusted_intercept = model_intercept + model_slope * min(Year)`


### Deriving min and max years (x-axis limits for trend lines)

```{r}
# !------ Use metadata from Part 1 file to create list containing all non NaN values by variable with 10+ yrs of obs------!#
# Named vector for mapping metadata_column to var_col for use in function
metadata_to_varcol <- c(
  "nYears_Secchi" = "SECCHI_DEPTH",
  "nYears_CHLA" = "CHLa",
  "nYears_totNitrogen" = "NITROGEN_TOTAL",
  "nYears_totPhosphorus" = "PHOSPHORUS_TOTAL",
  "nYears_TSS" = "TSS",
  "nYears_DIN" = "DIN"
)

# Function to subset data based on a minimum years of observations threshold. All station-variables in `data` have already been filtered to include only those with min 3 measurements per year (May-Oct).
subset_variable_data <- function(metadata_column, threshold = 10) {
  stations <- metadata %>%
    filter(!!sym(metadata_column) >= threshold) %>%
    pull(STATION_ID)
  
  var_col <- metadata_to_varcol[metadata_column]
    
  return(data %>%
           filter(STATION_ID %in% stations) %>%
           filter(!is.na(!!sym(var_col)))  
      )
}

# Create the subset_data list using the function
subset_data <- list(
  SECCHI_DEPTH = subset_variable_data("nYears_Secchi"),   
  CHLa = subset_variable_data("nYears_CHLA"),           
  NITROGEN_TOTAL = subset_variable_data("nYears_totNitrogen"), 
  PHOSPHORUS_TOTAL = subset_variable_data("nYears_totPhosphorus"), 
  TSS = subset_variable_data("nYears_TSS"),
  DIN = subset_variable_data("nYears_DIN")
)

```


```{r}
YearLims <- list()
vars <- c("SECCHI_DEPTH", "CHLa", "NITROGEN_TOTAL", "PHOSPHORUS_TOTAL", "TSS", "DIN")

for (var in vars) {
  # Subset the data for current var
  df <- subset_data[[var]]
  # Vector with unique stations for this variable
  the_stations <- unique(df$STATION_ID)
  
  # Initialize the data frame for storing minimum years
  minmax.df <- data.frame(STATION_ID = the_stations,
                           variable = var,
                           minYear = rep(NA, length(the_stations)),
                           maxYear = rep(NA, length(the_stations)))
  
  # Loop through the stations and pull the min & max year
  for (i in seq_along(the_stations)) {
    minmax.df$minYear[i] <- df %>% 
      filter(STATION_ID == the_stations[i]) %>%
      summarise(minYear = min(Year)) %>%
      pull(minYear)
    
    minmax.df$maxYear[i] <- df %>%
      filter(STATION_ID == the_stations[i]) %>%
      summarise(maxYear = max(Year)) %>%
      pull(maxYear)
  }
  
  # Store the data frame for this variable
  YearLims[[var]] <- minmax.df
}


yearLims.df <- bind_rows(YearLims)

```



```{r}
regressions_adjusted <- trends %>%
  left_join(yearLims.df, by = c("STATION_ID", "variable")) %>%
  mutate(adjusted_intercept = model_intercept + model_slope * minYear)

```

## Data Viz: mblm slopes for all stations, by variable

```{r}
# This code block is for setting up the data for plotting. It adds a field for significance and one with units using unicode characters.
plot_df <- regressions_adjusted %>%
  mutate(significant = if_else(model_pval <= .05 & model_slope > 0, "significant positive", 
                               if_else(model_pval <= .05 & model_slope < 0, "significant negative", "not significant"))
         ) %>%
  arrange(factor(significant, levels = c("not significant", "significant negative", "significant positive"))) %>% # set significant as factor, then arrange to ensure sig lines sit on top of insignificant lines
  mutate(var_units = case_when(
    variable == "SECCHI_DEPTH" ~ "Secchi Depth (m)",  # meters
    variable == "CHLa" ~ "Chlorophyll-a (\u03BCg L\u207B\u00B9)",  # micrograms per liter
    variable == "TSS" ~ "TSS (mg L\u207B\u00B9)",  # milligrams per liter
    variable == "NITROGEN_TOTAL" ~ "Total N (mg N L\u207B\u00B9)",  # milligrams of nitrogen per liter
    variable == "PHOSPHORUS_TOTAL" ~ "Total P (mg P L\u207B\u00B9)",  # milligrams of phosphorus per liter
    variable == "DIN" ~ "DIN (mg N L\u207B\u00B9)"  # milligrams of nitrogen per liter
  ))

# Remove specific outlier stations
plot_df <- plot_df %>%
  filter(!STATION_ID %in% c("2-XDD000.40", "5ASRN000.66"))

```


The approach used to plot (i.e., facet grid) in the previous publication will not work effectively here. In that instance we were facetting based on two variables: temp/do and then surface/bottom/mean/range. The x axis was the same for all plots, and the y axis was the same on each row. Here we have 6 varaibles each with different units, scales, and year ranges. My approach is to create 6 separate plots and then patch them together.

```{r fig.width=7, fig.height=7}
library(ggplot2)
library(patchwork)  

# Plot 1: Secchi Depth
p1 <- ggplot(subset(plot_df, variable == "SECCHI_DEPTH")) +
  geom_segment(aes(x = minYear, xend = maxYear, 
                   y = adjusted_intercept, 
                   yend = adjusted_intercept + model_slope * (maxYear - minYear), 
                   color = significant), size = .24) +
  scale_color_manual(values = c("not significant" = "grey72", 
                                "significant positive" = "#C9A818", 
                                "significant negative" = "#124E78")) +
  theme_minimal() + 
  labs(x = NULL, y = "Secchi Depth (m)") +  # Custom y-axis label
  theme(legend.position = "none",
        panel.border = element_rect(colour = "black", fill=NA, size=.5),
        axis.title.y = element_text(size = 10.2), 
        axis.text = element_text(color = "black"),
        axis.text.y = element_text(size = 8.5,  margin = margin(t = 0, r = 3, b = 0, l = 0)), 
        axis.text.x = element_text(size = 8.5, angle = 45, hjust = .95),
        legend.text = element_text(size = 8),  
        legend.title = element_blank(),
        strip.text = element_text(size = 11),
        strip.placement = "outside",
        panel.grid = element_line(color = "grey90", linewidth = .1),
        aspect.ratio = 1.8)

# Plot 2: Chlorophyll-a
#!---- This plot controls the LEGEND ----!#
p2 <- ggplot(subset(plot_df, variable == "CHLa")) +
  geom_segment(aes(x = minYear, xend = maxYear, 
                   y = adjusted_intercept, 
                   yend = adjusted_intercept + model_slope * (maxYear - minYear), 
                   color = significant), size = .24) +
  scale_color_manual(values = c("not significant" = "grey72", 
                                "significant positive" = "#C9A818", 
                                "significant negative" = "#124E78")) +
  theme_minimal() + 
  labs(x = NULL, y = "Chlorophyll-a (µg L\u207B\u00B9)") +  
  theme(legend.position = "top",
        legend.key.height = unit(1, "cm"),
        panel.border = element_rect(colour = "black", fill=NA, size=.5),
        axis.title.y = element_text(size = 10.2), 
        axis.text = element_text(color = "black"),
        axis.text.y = element_text(size = 8.5,  margin = margin(t = 0, r = 3, b = 0, l = 0)), 
        axis.text.x = element_text(size = 8.5, angle = 45, hjust = .95),
        legend.text = element_text(size = 9),  
        legend.title = element_blank(),
        strip.text = element_text(size = 11),
        strip.placement = "outside",
        panel.grid = element_line(color = "grey90", linewidth = .1),
        aspect.ratio = 1.8) +
  guides(color = guide_legend(override.aes = list(linewidth = 1.25, alpha = .9))) #adjust the width of the legend lines

# Plot 3: Total Suspended Solids (TSS)
p3 <- ggplot(subset(plot_df, variable == "TSS")) +
  geom_segment(aes(x = minYear, xend = maxYear, 
                   y = adjusted_intercept, 
                   yend = adjusted_intercept + model_slope * (maxYear - minYear), 
                   color = significant), size = .24) +
  scale_color_manual(values = c("not significant" = "grey72", 
                                "significant positive" = "#C9A818", 
                                "significant negative" = "#124E78")) +
  theme_minimal() + 
  labs(x = NULL, y = "TSS (mg L\u207B\u00B9)") +  
  theme(legend.position = "none",
        panel.border = element_rect(colour = "black", fill=NA, size=.5),
        axis.title.y = element_text(size = 10.2), 
        axis.text = element_text(color = "black"),
        axis.text.y = element_text(size = 8.5,  margin = margin(t = 0, r = 3, b = 0, l = 0)), 
        axis.text.x = element_text(size = 8.5, angle = 45, hjust = .95),
        legend.text = element_text(size = 8),  
        legend.title = element_blank(),
        strip.text = element_text(size = 11),
        strip.placement = "outside",
        panel.grid = element_line(color = "grey90", linewidth = .1),
        aspect.ratio = 1.8)


# Plot 4: Total Phosphorus
#! ---- This plot controls the X-AXIS LABEL (YEAR) ----!#
p4 <- ggplot(subset(plot_df, variable == "PHOSPHORUS_TOTAL")) +
  geom_segment(aes(x = minYear, xend = maxYear, 
                   y = adjusted_intercept, 
                   yend = adjusted_intercept + model_slope * (maxYear - minYear), 
                   color = significant), size = .24) +
  scale_color_manual(values = c("not significant" = "grey72", 
                                "significant positive" = "#C9A818", 
                                "significant negative" = "#124E78")) +
  theme_minimal() + 
  labs(x = NULL, y = "Total P (mg P L\u207B\u00B9)") +  
  theme(legend.position = "none",
        panel.border = element_rect(colour = "black", fill=NA, size=.5),
        axis.title.x = element_text(size = 11.3), 
        axis.text = element_text(color = "black"),    
        axis.title.y = element_text(size = 10.2),
        axis.text.y = element_text(size = 8.5,  margin = margin(t = 0, r = 3, b = 0, l = 0)), 
        axis.text.x = element_text(size = 8.5, angle = 45, hjust = .95),
        legend.text = element_text(size = 8),  
        legend.title = element_blank(),
        strip.text = element_text(size = 11),
        strip.placement = "outside",
        panel.grid = element_line(color = "grey90", linewidth = .1),
        aspect.ratio = 1.8)

# Plot 5: Total Nitrogen
#! ---- This plot controls the X-AXIS LABEL (YEAR) ----!#
p5 <- ggplot(subset(plot_df, variable == "NITROGEN_TOTAL")) +
  geom_segment(aes(x = minYear, xend = maxYear, 
                   y = adjusted_intercept, 
                   yend = adjusted_intercept + model_slope * (maxYear - minYear), 
                   color = significant), size = .24) +
  scale_color_manual(values = c("not significant" = "grey72", 
                                "significant positive" = "#C9A818", 
                                "significant negative" = "#124E78")) +
  scale_y_continuous(limits = c(0, 2.5), breaks = c(0, 1, 2)) +
  scale_x_continuous(limits = c(1990, 2025), breaks = seq(1990, 2020, 10)) +
  theme_minimal() + 
  labs(x = NULL, y = "Total N (mg N L\u207B\u00B9)") +  
  theme(legend.position = "none",
        panel.border = element_rect(colour = "black", fill=NA, size=.5),
        axis.title.y = element_text(size = 10.2), 
        axis.text = element_text(color = "black"),       
        axis.text.y = element_text(size = 8.5,  margin = margin(t = 0, r = 3, b = 0, l = 0)), 
        axis.text.x = element_text(size = 8.5, angle = 45, hjust = .95),
        legend.text = element_text(size = 8),  
        legend.title = element_blank(),
        strip.text = element_text(size = 11),
        strip.placement = "outside",
        panel.grid = element_line(color = "grey90", linewidth = .1),
        aspect.ratio = 1.8)


# Plot 6: Dissolved Inorganic Nitrogen (DIN)
p6 <- ggplot(subset(plot_df, variable == "DIN")) +
  geom_segment(aes(x = minYear, xend = maxYear, 
                   y = adjusted_intercept, 
                   yend = adjusted_intercept + model_slope * (maxYear - minYear), 
                   color = significant), size = .24) +
  scale_color_manual(values = c("not significant" = "grey72", 
                                "significant positive" = "#C9A818", 
                                "significant negative" = "#124E78")) +
  scale_y_continuous(limits = c(0, 2.5), breaks = c(0, 1, 2)) +
  scale_x_continuous(limits = c(1972, 2025), breaks = seq(1980, 2020, 10)) +
  theme_minimal() + 
  labs(x = NULL, y = "DIN (mg N L\u207B\u00B9)") +  # Custom y-axis label
  theme(legend.position = "none",
        panel.border = element_rect(colour = "black", fill=NA, size=.5),
        axis.title.y = element_text(size = 10.2), 
        axis.text = element_text(color = "black"),        
        axis.text.y = element_text(size = 8.5,  margin = margin(t = 0, r = 3, b = 0, l = 0)), 
        axis.text.x = element_text(size = 8.5, angle = 45, hjust = .95),
        legend.text = element_text(size = 8),  
        legend.title = element_blank(),
        strip.text = element_text(size = 11),
        strip.placement = "outside",
        panel.grid = element_line(color = "grey90", linewidth = .1),
        aspect.ratio = 1.8)

```

```{r fig.width=7, fig.height=7, warning = FALSE}
# Combine the six individual plots using patchwork
combined_plot <- (p1 | p2 | p3) /
                 (p4 | p5 | p6) 

# Display the combined plot
combined_plot

ggsave("figures/stationLevelTrends_MayOctavg.jpg", width = 7, height = 7, units = "in", dpi = 1000)



```

## Summary Statistics for Plot Stations

Produce a file that contains the long-term mean and median values for stations included in the plot, i.e., the average of the yearly May-Oct averages for Secchi, CHLa, TSS, TN, TP and DIN. 

```{r}
stations <- unique(plot_df$STATION_ID)

# Derive May-Oct Means for each year by station-variable
stationAvgs <- list()

for (var in names(subset_data)) {
  df <- subset_data[[var]]
  
  # Group by Year and STATION_ID, and calculate the May-Oct average for the variable
  sy_avgs <- df %>%
    group_by(Year, STATION_ID) %>%
    summarise(MAY_OCT_AVG = mean(!!sym(var), na.rm = TRUE))
  
  stationAvgs[[var]] <- sy_avgs
}

# Combine the station averages into a single data frame
stationAvgs.df <- bind_rows(stationAvgs, .id = "variable") %>%
  filter(STATION_ID %in% stations) %>%
  group_by(STATION_ID, variable) %>%
  summarise(mean = mean(MAY_OCT_AVG, na.rm = TRUE),
            median = median(MAY_OCT_AVG, na.rm = TRUE))

write_csv(stationAvgs.df, "data/stationAverages_longTerm.csv")

```



